---
title: "Extensions to the Grammar of Graphics"
subtitle: "Package 'ggpp' `r packageVersion('ggpp')` "
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Extensions to the Grammar of Graphics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
eval_all <- FALSE  # if TRUE evaluate all code chunks
```

## Preliminaries

We load all the packages used in the examples, if installed, and set
flags to skip examples if they are not available.

```{r, message=FALSE}
library(ggpp)
library(tibble)
library(dplyr)
eval_ggrepel <- requireNamespace("ggrepel", quietly = TRUE) &&
  packageVersion("ggrepel") > "0.9.1"
if (eval_ggrepel) library(ggrepel)
eval_magick <- requireNamespace("magick", quietly = TRUE)
```

As we will use text and labels on the plotting area we change the
default theme to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

# Introduction to the extensions

## Data labels, annotations and insets

*Data labels* add textual information directly related to individual
data points (shown as glyphs). Text position in this case is dependent
on the scales used to represent data points. Text is usually displaced
so that it does not occlude the glyph representing the data point and
when the link to the data point is unclear, this link is signaled with a
line segment or arrow. Data labels are distinct from annotations but
instead contribute to the representation of data on a plot or map.

*References* are lines, shading or marks used to help the reading of a
plot. These elements are used to highlight specific values on an axis or
a region in a plot. They, like data labels, are positioned relative to
the scales used for data. The position of data labels and lines, glyphs
or shading used as reference cannot be altered by the designer of a
plot, as the position conveys information. I will use the term *data
labels* irrespective if the "labels" are textual or graphical, like
icons and small plots and simple tables linked to data points or map
features.

According to Koponen and Hildén (2019), in a statistical chart 
"annotations can be used to draw reader attention to relevant detail".
These authors use as an example a text box in a plot to highlight a data
point that is off-scale and has been "squeezed" to a position
immediately outside the plotting area.

*Annotations* differ from data labels, in that their position is
decoupled from their meaning. Insets can be thought as larger, but still
self-contained annotations. In most cases the reading of inset tables
and plots depends only weakly on the plot or map in which they are
included.

In the case of annotations and insets the designer of a data
visualization has the freedom to locate them anywhere, as long they do
not occlude features used to describe data. I will use the term
*annotation* irrespective if the "labels" are textual or graphical.
Insets are similar to annotations, but the term *inset* is used when an
annotation's graphical or textual element is complex and occupies more
space within the plotting area. Insets can be moved from within the main
plotting area to being adjacent to it, e.g., as a smaller panel, without
any loss of meaning .

That the position of annotations and insets is independent of the
plotted data cannot be expressed using the grammar of graphics (GG) as
implemented in package 'ggplot2'.

## Annotations in 'ggplot2'

The plotting of data using package 'ggplot2' is described by Wickham
(2016) and several other books. Chapter 7 in Aphalo (2020) not only
describes the grammar used by 'ggplot2' but also several extensions to
it, including those provided by packages 'ggpp' and 'ggpmisc'.

In *The Layered Grammar of Graphics* (Wickham 2000) as implemented in
package 'ggplot2' (Wickham 2016) annotations are "second class"
features. As layers they behave differently than data layers: Only
constant values can be mapped to aesthetics and do no support faceting
into panels. In essence `annotate()` disconnects the resulting plot
elements from the data source and faceting, but not from the scales used
to graphically display the data values.

From the data visualization perspective the main practical and
conceptual difference between data labels and annotations is in the
scales used to position them within the plotting area. Instead of using
an `annotate()` function that deviates from the grammar of graphics to
implement annotations, we could retain the use of the grammar of
graphics for annotations but add support of native plot coordinates
(npc). Support of annotations done in this way would allow "annotation
layers" to behave almost identically to "data layers", and use the same
grammar. The *x* and *y* position aesthetics used for data could be
supplemented with pseudo-aesthetics without any translation relative to
the native plotting coordinates of the plotting area or viewport. Doing
so, would allow the graphic design flexibility conceptually inherent to
annotations within a user-friendly syntax.

## Extending 'ggplot2'

Based on this insight, a new approach to adding annotations and insets
was implemented in package 'ggpmisc' (>= 0.3.1) through two new *x* and
*y* pseudo-aesthetics, *npcx* and *npcy*, and corresponding dumb scales
and various geometries that make use of them. These scales and
geometries were implemented as an extension to package 'ggplot2' and are
in package 'ggpp'.

Before showing how this works in practice, we load the package and set
an uncluttered theme as default.

```{r, message=FALSE}
library(ggpp)
theme_set(theme_bw() + theme(panel.grid = element_blank()))
```

Native plot coordinates have range 0..1. However, the "npc" geoms also
recognize some positions by name with the same character strings as used
for text justification in 'ggplot2'. The default justification is
`"inward"`, as this protects from clipping at the edges the plotting
area irrespective the arguments passed to the `witdth` and `height`
parameters of R's graphic devices.

```{r}
p1 <- ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(x = c("left", "left"),
                                  y = c("top", "bottom"),
                                  label = c("Most\nefficient",
                                            "Least\nefficient")),
                mapping = aes(npcx = x, npcy = y, label = label),
                size = 3)
p1

```

The advantage of this approach becomes apparent when the limits of the
*y* scale are unknown or vary. When a script or user defined plotting
function sets the scale limits based on the input data, in the absence
of the extensions proposed here, setting annotations consistently within
the plotting area becomes laborious. The example below, shows how the
annotations remain at the desired position when the *y* scale limits are
expanded.

```{r}
p1 + expand_limits(y = 0)
```

To support the existing syntax for annotations using the new geometries,
function `ggplot2::annotate()` is overridden when package 'ggpp' is
loaded. The new definition adds support for the new pseudo-aesthetics
*npcx* and *npcy* retaining its original 'ggplot2' behaviour in all
other respects.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  annotate(geom = "text_npc",
           npcx = c("left", "left"),
           npcy = c("top", "bottom"),
           label = c("Most\nefficient",
                     "Least\nefficient"),
           size = 3)
```

Inset plots can be added with the same syntax using `geom_plot_npc()`.
They can be thought also as being annotations. Here we use `annotate()`
but `geom_plot()` can be also used directly, in which case the inset
plots can be different for each panel.

```{r}
p2 <- ggplot(mtcars, aes(factor(cyl), mpg, colour = factor(cyl))) +
  stat_boxplot() +
  labs(y = NULL) +
  theme_bw(9) + 
  theme(legend.position = "none",
        panel.grid = element_blank())

ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point() +
  annotate("plot_npc", npcx = "left", npcy = "bottom", label = p2) +
  expand_limits(y = 0, x = 0)

```

A simple example with facets, labelling of panels in a traditional way
as required by some book and journal styles. In this case panel tags are
added within the plotting area at a consistent "npc" location with free
scale limits in panels.

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(cyl = levels(factor(mtcars$cyl)),
                                  label = LETTERS[seq_along(levels(factor(mtcars$cyl)))],
                                  x = 0.90,
                                  y = 0.95),
                mapping = aes(npcx = x, npcy = y, label = label),
                size = 4) +
  facet_wrap(~factor(cyl), scales = "free") +
  theme(strip.background = element_blank(),
        strip.text = element_blank())
```

This approach was first implemented in 'ggpmisc' version 0.3.1 released
in April 2919. The implementation now in package 'ggpp' can be
considered stable. However, this implementation is to an extent
dependent on undocumented behaviour of 'ggplot2' functions, which means
that future updates to 'ggplot2' could break this functionality.

## When to use annotations and when insets

Package 'ggpmisc' adds support for various plot annotations and
reference guides based on model fits and other statistics. It also adds
support for some data labels related to data features. While developing
these statistics in 'ggpmisc', it became clear that expanding the
grammar of graphics's support for annotations would simplify the new
code considerably and also more cleanly separate the computations on
data from the positioning of annotations. These extensions to the
grammar are now in 'ggpp'.

**Table**. Geometries useful for data labels and annotations. Currently
implemented *ordinary* geometries and their *npc* versions. The
rightmost column shows the expected class of the objects mapped to the
`label` aesthetic. When using `annotate()` to add a single plot, table
or grob as an inset, enclosing them in a list is allowed, but not a
requirement. All those geoms whose names are highlighted in *`italics`*
support plotting of connecting segments or arrows when a the coordinates
have been modified by a *position* function.

+-------------------------------+---------------------+----------------------+
| Data labels                   | Annotations  (npc)  | `label` aes.         |
+===============================+=====================+======================+
| `ggplot2::geom_text()`        | `geom_text_npc()`   | `character`          |
+-------------------------------+---------------------+----------------------+
| `ggplot2::geom_label()`       | `geom_label_npc()`  | `character`          |
+-------------------------------+---------------------+----------------------+
| `ggrepel::geom_text_repel()`  |                     | `character`          |
+-------------------------------+---------------------+----------------------+
| `ggrepel::geom_label_repel()` |                     | `character`          |
+-------------------------------+---------------------+----------------------+
| `geom_text_s()`               |                     | `character`          |
+-------------------------------+---------------------+----------------------+
| `geom_label_s()`              |                     | `character`          |
+-------------------------------+---------------------+----------------------+
| `geom_point_s()`              |                     |                      |
+-------------------------------+---------------------+----------------------+
| `geom_plot()`                 | `geom_plot_npc()`   | `list(<ggplot>)`     |
+-------------------------------+---------------------+----------------------+
| `geom_table()`                | `geom_table_npc()`  | `list(<data.frame>)` |
+-------------------------------+---------------------+----------------------+
| `geom_grob()`                 | `geom_grob_npc()`   | `list(<grob>)`       |
+-------------------------------+---------------------+----------------------+
| `geom_img()`  *(planned)*     | `geom_img_npc()`    | `list(<raster>)`     |
+-------------------------------+---------------------+----------------------+

: Geometries

When adding an informative element to a plot, assess whether it is an
*annotation* or a *data label*. To decide on the best approach, consider
if the location of the element is more "naturally" expressed in the
original data units or as position relative to the edges or centre of
the plotting area. In the second case, prefer the "npc" geoms as you are
dealing with *annotations*, otherwise, use the ordinary geometries as
you are dealing with data labels or data points.

## Positioning data labels

Nudging is normally used to displace text labels so that they do not
overlap the points or lines representing the data being labelled. Two
enhanced versions of `position_nudge()` are provided,
`position_nudge_center()` and `position_nudge_line()`. These functions
make it possible to apply nudging that varies automatically according to
the relative position of points with respect to arbitrary points or
lines, or with respect to a polynomial or smoothing spline fitted
on-the-fly to the the observations.

A limitation of 'ggplot2' is that only one position function can be used
in a layer. This makes it difficult to add text labels to stacked or
dodged bars or columns, or the labelling of points that have been
jittered. The solution provided by 'ggpp' are wrappers on these position
functions that add support for nudging.

All these new position functions, instead of deleting the original
positions, they rename the variables in `data` containing these
coordinates. This makes it possible for geometries to draw connecting
segments or arrows between old and new positions. These functions are
backwards compatible and can be used in place of those from 'ggplot2'.
The geometries from 'ggrepel' (> 0.9.1) and many of those from 'ggpp'
can make use of the original coordinates kept in data.

| Position (origin kept)                      | Position (origin discarded)       |
|---------------------------------------------|-----------------------------------|
| `position_nudge_keep()`                     | `ggplot2::position_nudge()`       |
| `ggrepel::postion_nudge_repel()`            | `ggplot2::position_nudge()`       |
| `position_nudge_centre()`                   |                                   |
| `position_nudge_line()`                     |                                   |
| `position_jitter_keep()`                    | `ggplot2::position_jitter()`      |
| `position_dodge_keep()`                     | `ggplot2::position_dodge()`       |
| `position_dodge2_keep()`                    | `ggplot2::position_dodge2()`      |
| `position_stack_keep()`                     | `ggplot2::position_stack()`       |
| `position_fill_keep()`                      | `ggplot2::position_fill()`        |
| `position_jitterdodge_keep()` *(planned)*   | `ggplot2::position_jitterdodge()` |
| `position_jitternudge()`                    |                                   |
| `position_dodgenudge()`                     |                                   |
| `position_dodge2nudge()`                    |                                   |
| `position_stacknudge()`                     |                                   |
| `position_fillnudge()`                      |                                   |
| `position_jitterdodgenudge()` *(planned)*   |                                   |

: Position functions

## Highlighting data features

If maxima or minima are to be labelled in a plot, this can be automated
so that the feature and its labelling is automated. This is possible
with `stat_peaks()` and `stat_valleys()`. Marginal annotations with
symbols and vertical and horizontal lines can be used to highlight
summary values or events/conditions. The *geoms*
`geom_quadrant_lines()`, `geom_vhlines()`, `geom_x_margin_arrow()`,
`geom_x_margin_grob()`, `geom_x_margin_point()`,
`geom_y_margin_arrow()`, `geom_y_margin_grob()` and
`geom_y_margin_point()` make such labelling easy.

## Local density of observations

There are two aspects to this: highlighting with colours dense or sparse
regions, or using density to decide which observations to label. In this
cases we can use `stat_dens1d_labels()` and `stat_dens2d_labels()`,
`stat_dens1d_filter()` and `stat_dens2d_filter()`, and
`stat_dens1d_filter_g()` and `stat_dens2d_filter_g()`.

## Summary statistics

With some types of data it is common to compute a summary per quadrant,
rather than based on a prior grouping. The *stat*
`stat_quadrant_counts()` can de used to annotate plots with the number
of observations in each quadrant. `stat_summary_xy()` is similar to
`stat_summary()` from 'ggplot2' but useful when both *x* and *y* are
continuous variables. It allows the use the same, or different functions
to summarise the data over *x* and *y*. In contrast `stat_centroid()`
always applies the same function over *x* and *y*.

## ggplot methods

Being `ggplot()` defined as a generic method in 'ggplot2' makes it
possible to define specializations, and we provide two for time series
stored in objects of classes `ts` and `xts` which automatically convert
these objects into tibbles and set by default the aesthetic mappings for
`x` and `y` automatically. A companion function `try_tibble()` is also
exported.

## Thoughts for the future

Although we have discussed data independent scales only for *x* and *y*
aesthetics, the same consideration applies to other aesthetics like
*colour*. It seems generally useful to allow aesthetics to use a
different scale for annotations than for data. Defining
pseudo-aesthetics for annotations instead of allowing multiple scales
for each aesthetic in a plot would add flexibility while still keeping
the guarantee that the meaning of aesthetics' values remains consistent
across all plot elements representing data.

## List of references

Aphalo, Pedro J. (2020) *Learn R: As a Language.* The R Series. Boca
Raton and London: Chapman and Hall/CRC Press. ISBN: 978-0-367-18253-3.
350 pp.

Koponen, J; Hildén, J. (2019) *Data visualization handbook*. Aalto ARTS
books, Espoo. ISBN 978-952-60-7449-8.

Wickham H. (2010) A Layered Grammar of Graphics. *Journal of
Computational and Graphical Statistics* **19:** 3--28.

Wickham H. (2016) *ggplot2: Elegant Graphics for Data Analysis*.
Springer International Publishing. ISBN 978-3-319-24275-0.

# Examples

Some of the examples below are evaluated only if packages 'ggrepel', and
'magick' are available.

## ggplot methods

`ggplot()` methods for classes `"ts"` and `"xts"` automate plotting of
time series data, as *x* and *y* aesthetics are mapped to time and the
variable of the time series, respectively. For plotting time series data
stored in objects of other classes, see the conversion functions
`try_tibble()` and `try_data_frame()` in the last section of this
vignette.

By default, `time` is of class `Date` or of class `POSIXct` depending on
how time is stored in the time series object. On-the-fly coversion of
`time` to `numeric` is also possible (shown in the second example
below).

```{r}
class(lynx)
ggplot(lynx) + geom_line()
```

The class of variable `time`, mapped to the *x* aesthetic, affects the
scale used by default as well as the formatting of values when converted
to character strings or printed. Here we force its conversion to
`numeric`.

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line()
```

## Geometries

Three of the geometries described below allow the addition of plot
layers containing insets. Insets can be plots, tables, bitmaps, or grid
objects. Insets can be also added as annotations. Using for `data` a
tibble with a list column containing data frames or tibbles allows like
any other geom, the use of grouping, multiple insets per panel, faceting
with different tables per panel, and different number of insets in each
panel, i.e., individual tables added to a plot with `geom_table` behave
similarly to individual `character` values added with `geom_text`.

Other geometries also described in this section support the use of
native plot coordinates for positioning elements in the plotting area.
Obviously these geometries are not meant to be used to display data, but
instead they make it possible to add annotations to plots consistently
across data sets even when using scales with varying limits.

### geom_table() and stat_fmt_table()

The *geometry* `geom_table()` plots a data frame or tibble, nested in a
tibble passed as data argument, using *aesthetics* `x` and `y` for
positioning, and `label` for the list of data frames containing the data
for the tables. The tables are created as 'grid' grobs and added as
usual to the ggplot object. In contrast to "standard" geoms, this geom
by default does not inherit the globally mapped aesthetics. Tables are
always added at their native sizes, which can be altered by changing the
`size` of the text in them.

```{r}
tb <- mpg %>%
  group_by(cyl) %>%
  summarise(hwy = median(hwy), cty = median(cty))

data.tb <- tibble(x = 7, y = 44, tb = list(tb))

ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_table(data = data.tb, aes(x, y, label = tb)) +
  geom_point() 
```

In plots with a single panel it can be easier to use `annotate()` to add
inset tables, giving the same plot as above. In this case single data
frames, ggplots or grobs do not need to be wrapped in a list, although
lists are also supported.

```{r, eval=eval_all}
tb <- mpg %>%
  group_by(cyl) %>%
  summarise(hwy = median(hwy), cty = median(cty))

ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  annotate("table", x = 7, y = 44, label = tb) +
  geom_point() 
```

Table themes are supported through parameter `table.theme` and if
variables or constants are mapped to the `colour`, `fill`, `size`, or
`family`, aesthetics they override the corresponding default theme
settings. The display of `rownames` and `colnames` can be enabled or
disable through parameter `table.rownames` and `table.colnames` and the
horizontal justification of text in the core of the table through
parameter `table.hjust`.

Parameter `table.theme` accepts as arguments `NULL` for use of the
current default, a `ttheme` constructor function such as those defined
in package 'gridExtra', or the variations on them defined in this
package. The active default can be set with function `ttheme_set()`.

```{r}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_table(data = data.tb, aes(x, y, label = tb),
             table.theme = ttheme_gtsimple,
             table.hjust = 0, colour = "darkred", fill = "#FFFFBB") +
  geom_point() 
```

Using `stat_fmt_tb()` we can rename columns and rows of the tibble,
reorder them and/or select a subset of columns or rows as shown below.
To provide a complete example we also replace the names of the scales
for *x*, *y* and *color* aesthetics. Here we pass a `character` vector
with the original names of the columns in full, but partial matching is
tried when needed. It is also possible to use a `numeric` vector of
positional indexes.

```{r}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_table(data = data.tb, aes(x, y, label = tb),
             table.theme = ttheme_gtlight,
             size = 3, colour = "darkblue",
             stat = "fmt_tb", 
             tb.vars = c(Cylinders = "cyl", MPG = "hwy"), # rename
             tb.rows = 4:1) + # change order
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine cylinders\n(number)") +
  geom_point() +
  theme_bw()
```

Parsed text, using plot math syntax is supported in the table, with
fall-back to plain text in case of parsing errors, on a cell by cell
basis (see `tableGrob()` in package 'gridExtra' for details, as this
function is used to build the table). Here we plot the MPG for city
traffic and we can see that the plotting area expands to include the
coordinates at which the table is anchored. Justification is by default
set to `"inward"` which ensures that the table is fully within the
plotting region.

```{r}
tb.pm <- tibble(Parameter = c("frac(beta[1], a^2)", "frac(beta[2], a^3)"),
                Value = c("10^2.4", "10^3.532"))
data.tb <- tibble(x = 7, y = 44, tb = list(tb.pm))
ggplot(mpg, aes(displ, cty)) +
  geom_point() +
  geom_table(data = data.tb, aes(x, y, label = tb), parse = TRUE) +
  theme_bw()
```

As implemented, there is no limitation to the number of insets, and
faceting is respected. If the base plot shows a map, multiple small
tables could be superimposed on different countries or regions. The size
of the table inset is given by the *size* aesthetic, so like for
`geom_text()` it is independent of the `ggplot2::theme()` and controlled
by the `table.theme` directly or indirectly through the *size*
aesthetic.

Please see section **Normalised Parent Coordinates** below for a
description of `geom_table_npc()`.

### geom_plot()

The `geom_plot()` *geometry* plots ggplot objects, nested in a tibble
passed as data argument, using *aesthetics* `x` and `y` for positioning,
and `label` for the ggplot object containing the definition of the plot
to be nested. With this approach in plots with facets the insets can be
different in each panel. It is also possible to inset more than one plot
in a single call simply by creating a tibble with multiple rows.

Behind the scenes, one Grob is created for each plot to be inset. The
conversion is done with `ggplotGrob()` and the Grobs added to the main
ggplot object.

As an example we produce a plot where the inset plot is a zoomed-in
detail from the main plot. In this case the main and inset plots start
as the same plot. In most cases the size of text and other elements in
the inset should be smaller than in the main plot. Here we override the
default theme setting the `base_size` from its default of 11 pt to 8 pt.

```{r}
p <- ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() 

data.tb <- 
  tibble(x = 7, y = 44, 
         plot = list(p + 
                       coord_cartesian(xlim = c(4.9, 6.2), 
                                       ylim = c(13, 21)) +
                       labs(x = NULL, y = NULL) +
                       theme_bw(8) +
                       scale_colour_discrete(guide = "none")))

ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_plot(data = data.tb, aes(x, y, label = plot)) +
  annotate(geom = "rect", 
           xmin = 4.9, xmax = 6.2, ymin = 13, ymax = 21,
           linetype = "dotted", fill = NA, colour = "black") +
  geom_point() 
```

In general, the inset plot can be any ggplot object, allowing the
creation of very different combinations of main plot and inset plots.
Here we use the inset to show summaries as in the previous example of an
inset table.

```{r}
p <- ggplot(mpg, aes(factor(cyl), hwy, fill = factor(cyl))) +
  stat_summary(geom = "col", fun = mean, width = 2/3) +
  labs(x = "Number of cylinders", y = NULL, title = "Means") +
  scale_fill_discrete(guide = "none")

data.tb <- tibble(x = 7, y = 44, 
                  plot = list(p +
                                theme_bw(8)))

ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_plot(data = data.tb, aes(x, y, label = plot)) +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine cylinders\n(number)") +
  theme_bw()
```

The same plot as above can be created using `annotate()`, but be aware
that when using facets in 'ggplot2' annotations are identical in all
panels.

```{r, eval=eval_all}
p <- ggplot(mpg, aes(factor(cyl), hwy, fill = factor(cyl))) +
  stat_summary(geom = "col", fun = mean, width = 2/3) +
  labs(x = "Number of cylinders", y = NULL, title = "Means") +
  scale_fill_discrete(guide = "none")

ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  annotate("plot", x = 7, y = 44, label = p + theme_bw(8)) +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine cylinders\n(number)") +
  theme_bw()
```

As implemented, there is no limitation to the number of insets, and
faceting is respected. If the base plot shows a map or a bitmap,
multiple small plots could be superimposed on different countries or
regions. The size of the insets is controlled by the `vp.width` and
`vp.height` aesthetics as a fraction of the main plot's plotting region.
Consequently, the insets are scaled together with the main plot. A
possible unintuitive but useful feature, is that the theme is linked to
each plot.

Please see section **Normalised Parent Coordinates** below for a
description of `geom_plot_npc()`.

### geom_grob()

The `geom_grob()` *geometry* plots grobs (graphical objects as created
with 'grid'), nested in a tibble passed as data argument, using
*aesthetics* `x` and `y` for positioning, and `label` for the Grob
object. While `geom_table()` and `geom_plot()` take as values mapped to
the `label` aesthetics tibbles or data frames, and ggplots,
respectively, and convert them into Grobs, `geom_grob()` expects Grobs
ready to be rendered. This means that any Grob created using 'grid' or
its extenssions can be added as a data label to a ggplot.

```{r, eval=eval_magick}
file.name <- 
  system.file("extdata", "Isoquercitin.png", 
              package = "ggpp", mustWork = TRUE)
Isoquercitin <- magick::image_read(file.name)
grobs.tb <- tibble(x = c(0, 10, 20, 40), y = c(4, 5, 6, 9),
                   width = c(0.05, 0.05, 0.01, 1),
                   height =  c(0.05, 0.05, 0.01, 0.3),
                   grob = list(grid::circleGrob(), 
                               grid::rectGrob(), 
                               grid::textGrob("I am a Grob"),
                               grid::rasterGrob(image = Isoquercitin)))

ggplot() +
  geom_grob(data = grobs.tb, 
            aes(x, y, label = grob, vp.width = width, vp.height = height),
            hjust = 0.7, vjust = 0.55) +
  scale_y_continuous(expand = expansion(mult = 0.3, add = 0)) +
  scale_x_continuous(expand = expansion(mult = 0.2, add = 0)) +
  theme_bw(12)
```

As shown above for inset tables and inset plots, it is also possible to
use `annotate()` with Grobs. The next example insets a single `Grob`.
Here we reuse the bitmap `Isoquercitin` read in the previous example.
The Grob is contained in a viewport. Here setting `width = 1` ("npc"
units) when creating the `Grob` from the bitmap ensures that the bitmap
fills the width of the viewport (to ensure that the inset is not
distorted, set only one of `width` or `height`). The argument to
`vp.width` or `vp.height`, also in "npc" units, determines the size of
the `Grob` relative to the size of the plotting area.

```{r, eval=(eval_all && eval_magick)}
ggplot() +
  annotate("grob", x = 1, y = 3, vp.width = 0.5,
           label = grid::rasterGrob(image = Isoquercitin, width = 1)) +
  theme_bw(12)
```

`geom_grob()` is designed thinking that its main use will in graphical
annotations, although one could use it for infographics with multiple
copies of each grob, this would go against the grammar of graphics. In
this implementation grobs cannot be mapped to an aesthetic through a
scale.

As implemented, there is no limitation to the number of insets and
faceting is respected. If the base plot shows a map or a bitmap,
multiple simple grobs (e.g. national flags) could be superimposed on
different countries. The size of the insets is controlled by the
`vp.width` and `vp.height` aesthetics as a fraction of the main plot's
plotting region. Consequently, the insets are scaled together with the
main plot.

Please see section **Normalised Parent Coordinates** below for a
description of `geom_grob_npc()`.

### geom_vhlines()

This is a convenience geometry that adds both vertical and horizontal
guide lines on the same plot layer, using the same syntax as
`geom_hline()` and `geom_vline()` from package 'ggplot2'.

```{r}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_vhlines(xintercept = c(2.75, 4), yintercept = 27, linetype = "dashed") +
  geom_point() +
  labs(x = "Engine displacement (l)", y = "Fuel use efficiency (MPG)",
       colour = "Engine cylinders\n(number)") 

```

### geom_text_s() and geom_point_s()

With package 'ggrepel' we have become used to easily add text and labels
linked to the labelled observations or points by a line segment or
arrow. However, sometimes a plot has few text labels and repulsion is
unnecessary. In such cases nudging or dodging may be enough, or even
preferable so as to achieve consistency.

We can achieve this with `geom_text_s()` (formerly named
`geom_text_linked()`) together with a suitable position function that
retains the original *x* and *y* coordinates of the point. We currently
provide three different nudge functions described in a later section.
One of them is used by default providing the same default behaviour as
`geom_text()` from 'ggplot2' but drawing linking segments if nudging is
set to a value different to the default of zero. The equivalent for
points connected with segments is `geom_point_s()`.

```{r}
my.cars <- mtcars[c(TRUE, FALSE, FALSE, FALSE), ]
my.cars$name <- rownames(my.cars)
my.cars <- my.cars[order(my.cars$wt), ]
ggplot(my.cars, aes(wt, mpg, label = name)) +
  geom_point() +
  geom_text_s(aes(colour = factor(cyl)),
                   angle = 90,
                   hjust = 1.04, nudge_y = -1, 
                   nudge_x = c(0, 0, -0.05, +0.05, 0, 0 ,0, 0),
                   arrow = arrow(length = grid::unit(1.5, "mm"))) +
  scale_colour_discrete(l = 40) +
  expand_limits(y = 0)
```

### Normalised Parent Coordinates

R's 'grid' package defines several units that can be used to describe
the locations of plot elements. In 'ggplot2' the *x* and *y* aesthetics
are directly mapped to `"native"` or data units. For consistent location
of **annotations** with respect to the plotting area we need to rely on
`"npc"` which are expressed relative to the size of the grid viewport.
The plotting area in a ggplot is implemented as a 'grid' viewport and
support for `"npc"` coordinates is relatively easy to implement.

To support `"npc"` positions we have implemented *scales* for two new
(pseudo) *aesthetics*, `npcx` and `npcy`. These are very simple
continuous scales which do not support any transformation or changes to
their limits, both of which would be meaningless for `"npc"` units.
Variables mapped to these aesthetics can be either numerical with values
in the range zero to one or character. A limited set of strings are
recognised and converted to `"npc"` units: `"bottom"`, `"center"`,
`"top"`, `"left"`, `"middle"`, `"right"` (`"centre"` is a synonym for
`"center"`).

To make these scales useful we need also to define *geometries* that use
these new aesthetics. Package 'ggpp' currently provides
`geom_text_npc()`, `geom_label_npc()`, `geom_table_npc()`,
`geom_plot_npc()` and `geom_grob_npc()`.

As is the case for `geom_text()` and `geom_label()` from package
'ggplot2', `"bottom"`, `"center"`, `"top"`, `"left"`, `"middle"`,
`"right"`, plus `"inward"` and `"outward"` can be used, as well as
numeric values, to control the justification. Justification defaults to
`"inward"` in the geometries described here.

While the usual *x* and *y* aesthetics are used whenever the positions
of plot elements represent data values, these new scales and geometries
are useful only for annotations, i.e., in those cases when we want plot
elements at specific positions within the plotting area irrespective of
the ranges of the data mapped to the *x* and *y* aesthetics. When
writing scripts or functions that may be applied to different data sets
these new *aesthetics* help by keeping the code concise and reusable.
These *geometries* are used by default by several of the statistics
described in later sections and those defined in package 'ggpmisc'.

As an example let's imagine that we want to add the structure of a
metabolite to a plot. Its position has nothing to do with the data
mapped to *x* and *y*, so it is conceptually better to use `"npc"`
coordinates. The big practical advantage is that this also allows to
keep this part of the plot definition independent of the data being
plotted, giving a major advantage in the case of plots with facets with
free scale limits. This example can be easily adapted to
`geom_plot_npc()` where a list of ggplots is mapped to `label`, and to
`geom_table_npc()` where a list of data frame is mapped to `label`.

We produce the example plot by first constructing a tibble to contain
the grob and the coordinate data, and then map these variables to
aesthetics using `aes()`. In the example the tibble has a single row,
but this is not a requirement. In this respect these geoms behave as
normal geoms, with facets also supported.

```{r, eval=eval_magick}
file.name <- 
  system.file("extdata", "Robinin.png", 
              package = "ggpp", mustWork = TRUE)
Robinin <- magick::image_read(file.name)

set.seed(123456)
data.tb <- tibble(x = 1:20, y = (1:20) + rnorm(20, 0, 10))

flavo.tb <- tibble(x = 0.02,
                   y = 0.95,
                   width = 1/2,
                   height = 1/4,
                   grob = list(grid::rasterGrob(image = Robinin)))

ggplot(data.tb, aes(x, y)) +
  geom_grob_npc(data = flavo.tb, 
                aes(label = grob, npcx = x, npcy = y, 
                    vp.width = width, vp.height = height)) +
  geom_point() +
  expand_limits(y = 55, x = 0)
```

Alternatively, we can pass constant values to `geom_grob_npc()` to
obtain the same plot. This approach can be handy in simple cases.

```{r, eval=(eval_all && eval_magick)}
ggplot(data.tb, aes(x, y)) +
  geom_grob_npc(label = list(grid::rasterGrob(image = Robinin, width = 1)), 
                npcx = 0.02, npcy = 0.95,
                vp.width = 1/2, vp.height = 1/4) +
  geom_point() +
  expand_limits(y = 55, x = 0)
```

We can also use `annotate()` if the annotation should be the same for
all panels, or if we have a single figure panel. In this case there is
no need to wrap a single `grob` in a list.

```{r, eval=(eval_all && eval_magick)}
ggplot(data.tb, aes(x, y)) +
  annotate("grob_npc", label = grid::rasterGrob(image = Robinin, width = 1), 
                npcx = 0.02, npcy = 0.95, vp.width = 1/2, vp.height = 1/4) +
  geom_point() +
  expand_limits(y = 55, x = 0)

```

Two additional geometries are based on existing 'ggplot2' geometries.
They are based on `geom_text()` and `geom_label()`. We give an example
using `geom_text_npc()` to produce a "classic" labelling for facets
matching the style of `theme_classic()` and traditional scientific
journals' design.

```{r}
corner_letters.tb <- tibble(label = LETTERS[1:4],
                            x = "left", 
                            y = "top",
                            cyl = c(4,5,6,8))
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl, scales = "free") +
  geom_text_npc(data = corner_letters.tb,
                aes(npcx = x, npcy = y, label = label)) +
  theme_classic() +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank())
```

### Marginal markings

'ggplot2' provides `geom_rug()`, `geom_vline()` and `geom_hline()`. Rug
plots are intended to be used to represent distributions along the
margins of plot. `geom_vline()` and `geom_hline()`are normally used to
separate regions in a plot or to highlight important values along the
*x* or *y* axis. When creating plots it is sometimes useful to put small
marks along the axes, just inside the plotting area, similar to those in
a rug plot, but like `geom_vline()` and `geom_hline()` in their purpose.

Three *geometries* provide such markers: `geom_margin_point()`,
`geom_margin_arrow()`, and `geom_margin_grob()`. They behave similarly
to `geom_vline()` and `geom_hline()` and their positions are determined
also by the `xintercept` and `yintercept` aesthetics.

In the example below we indicate the group medians along the *x* axis
with filled triangles.

```{r}
data.tb <- mpg %>%
  group_by(cyl) %>%
  summarise(hwy = median(hwy), displ = median(displ))
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_x_margin_point(data = data.tb,
                      aes(xintercept = displ, fill = factor(cyl))) +
  expand_limits(y = 10) +
  geom_point() 
```

## Statistics

### stat_centroid()

It can be useful to mark the centroid of a group of observations with a
point or with a label. By default `stat_centroid()` applies function
`mean_se()` to both `x` and `y` by group. If the value mapped to an
aesthetics across rows within each group is unique this value is copied
to the returned `data`.

```{r}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point(alpha = 0.33) +
  stat_centroid(shape = "cross", size = 4)
```

Other functions can be passed to this statistic as long as they return a
single value that can be mapped to the `x` and `y` aesthetics (numeric,
time or a factor).

```{r, eval = FALSE}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point(alpha = 0.33) +
  stat_centroid(shape = "cross", size = 4, .fun = median)
```

The very similar `stat_summary_xy()` accepts different functions for `x`
and `y`.

### stat_quadrant_counts() and geom_quadrant_lines()

This statistic automates the annotation of plots with number of
observations, either by quadrant, by pairs of quadrants or the four
quadrants taken together (whole plotting area). Its companion geometry,
`geom_quadrant_lines()` is used in the examples to highlight the
quadrants.

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- -99:100
y <- x + rnorm(length(x), mean = 0, sd = abs(x))
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"))
```

Using defaults except for color.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_quadrant_lines(colour = "red") +
  stat_quadrant_counts(colour = "red") +
  geom_point() +
  expand_limits(y = c(-250, 250))
```

Pooling quadrants along the *x*-axis. (`pool.along = "y"` pools along
*y*.)

```{r, eval=eval_all}
ggplot(my.data, aes(x, y)) +
  geom_quadrant_lines(colour = "red", pool.along = "x") +
  stat_quadrant_counts(colour = "red", pool.along = "x") +
  geom_point()
```

Manual positioning of the text annotations and pooling of all four
quadrants, and overriding the default formatting for the label.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quadrant_counts(quadrants = 0L, label.x = "left", 
                       aes(label = sprintf("%i observations", stat(count))))
```

Annotation of only specific quadrants.

```{r, eval=eval_all}
ggplot(my.data, aes(x, y)) +
  geom_quadrant_lines(colour = "red") +
  stat_quadrant_counts(colour = "red", quadrants = c(2, 4)) +
  geom_point()
```

Using facets, even with free scale limits, the labels are placed
consistently. This achieved by the default use of `geom_text_npc()` or
as shown below by use of \`geom_label_npc(). We expand the *y* limits to
ensure that no observations are occluded by the labels.

```{r}
ggplot(my.data, aes(x, y, colour = group)) +
  geom_quadrant_lines() +
  stat_quadrant_counts(geom = "label_npc") +
  geom_point() +
  expand_limits(y = c(-260, 260)) +
  facet_wrap(~group)
```

### stat_apply_group

This statistic applies functions to `x` and `y` data. The function(s)
supplied as argument are expected to return a numeric vector. Both
functions should return vectors of the same length. When possible it is
usually better to apply functions through mappings using `aes()` when
they are independent of grouping, or to use a transformation for the
scale. However, when grouping is important, these statistics make it
possible to avoid pre-computation of the data for a layer.

Here we plot the difference in trunck circumference between dates for
each tree. (As the vector returned by `diff()` is one element shorter
than its input, we delete the first value of `x` in `.fun.x`.).

```{r}
ggplot(Orange, aes(age, circumference, colour = Tree)) +
  stat_apply_group(.fun.x = function(x) {x[-1L]},
                   .fun.y = diff)
```

### stat_dens2d_labels() and stat_dens2d_filter()

These stats had their origin in an enhancement suggestion for 'ggrepel'
from Hadley Wickham and discussion with Kamil Slowikowski (ggrepel's
author) and others. In fact the code is based on code Kamil gave during
the discussion, but simplified and taking a few further ideas from
`ggplot::stat_dens2d`.

**Warning!** Which observations are selected by the algorithm used,
based on `MASS:kde2d`, depends strongly on the values of parameters `h`
and `n`. You may need to alter the defaults by passing explicit
arguments to these stats. Beware, though, that what are good values, may
depend on individual data sets even if they include the same number of
observations. For the selection of observations to work cleanly, the
argument for `n` must create a dense grid. Much larger values of `n`
than in the examples in the documentation of `MASS::kde2d` and
`ggplot2::stat_dens2d` will be needed in most cases.

Some random data with random labels.

```{r}
random_string <- function(len = 3) {
paste(sample(letters, len, replace = TRUE), collapse = "")
}

# Make random data.
set.seed(1001)
d <- tibble::tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A", "B"), c(50, 50)),
  lab = replicate(100, { random_string() })
)
```

The stat `stat_dens2d_filter` *filters* observations, in other words
passes to the geom a subset of the data received as input. The default
argument for `geom` is `"point"`.

Using defaults except for the color aesthetic. Highlight 1/4 of
observations from lowest density areas of the plot panel.

```{r}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(keep.fraction = 1/4, colour = "red")
```

Keep at most 50 observations.

```{r, eval=eval_all}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(keep.fraction = 1/4, keep.number = 50, colour = "red")
```

Keep always 50 observations.

```{r, eval=eval_all}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(keep.fraction = 1, keep.number = 50, colour = "red")
```

Highlighting 1/4 of the observations by under-plotting with larger black
points.

```{r, eval=eval_all}
ggplot(data = d, aes(x, y, colour = group)) +
   stat_dens2d_filter(keep.fraction = 0.25,
                      size = 3,
                      colour = "black") +
   geom_point()
```

A different way of highlighting 1/4 of the observations, using
over-plotting with a 'hollow' shape. We also shift one group with
respect to the other.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, colour = group)) +
   geom_point() +
   stat_dens2d_filter(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

Highlight 1/4 of observations from lowest density areas of the plot,
with density considered separately for each individual group. In this
case grouping is based on the color aesthetic.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, colour = group)) +
   geom_point() +
   stat_dens2d_filter_g(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

The stat `stat_dens1d_labels` *replaces* the values of the label
(aesthetic) variable in `data` based on density of observations along
the *x* or *y* axis in the plot panel. The replacement is given by the
argument passed to `label.fill`, which can be a character string or a
function accepting a character string as argument and returning also a
character string.

The default value for `geom` is `"text"`. The default value of
`label.fill` is `""` which results in empty labels, while using `NA` as
fill label results in observations being omitted. Using `NA` as
`label.fill` is similar to using `stat_dens2d_filter` as long as the
geom used requires a `label` aesthetic.

Label 1/10 of observations from lowest density areas of the plot panels.

```{r, eval=eval_all}
ggplot(data = d, aes(x, y, label = lab, colour = group)) +
  stat_dens2d_labels(keep.fraction = 1/10, 
                     hjust = "outward", vjust = "outward") +
  geom_point()
```

Using the geoms from package 'ggrepel' avoids clashes among labels or on
top of data points. This works with versions 0.6.0 and newer of
'ggrepel'. One example with `geom_text_repel` follows.

```{r, eval=eval_ggrepel}
ggplot(data = d, aes(x, y, label = lab, colour = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "text_repel", 
                     keep.fraction = 0.45)
```

With `geom_label_repel` one usually needs to use a smaller value for
`keep.fracton`, or a smaller `size`, as labels use more space on the
plot than the test alone.

Additional arguments can be used to change the angle and position of the
text, but may give unexpected output when labels are long as the
repulsion algorithm "sees" always a rectangular bounding box that is not
rotated. With short labels or angles that are multiples of 90 degrees,
there is no such problem. Please, see the documentation for
`ggrepel::geom_text_repel` and `ggrepel::geom_label_repel` for the
various ways in which both repulsion and formatting of the labels can be
adjusted.

Using `NA` as argument to `label.fill` makes the observations with
labels set to `NA` *incomplete*, and such rows in data are skipped when
rendering the plot, before the repulsion algorithm is active. This can
lead to overlap between text and points corresponding to unlabelled
observations. Whether points are occluded depends on the order of layers
and transparency, the occlusion can remain easily unnoticed with
`geom_label` and `geom_label_repel`. We keep `geom_point` as the topmost
layer to ensure that all observations are visible.

```{r, eval=(eval_all && eval_ggrepel)}
ggplot(data = d, aes(x, y, label = lab, colour = group)) +
  stat_dens2d_labels(geom = "label_repel", 
                     keep.fraction = 0.2, 
                     label.fill = NA) +
    geom_point()
```

### stat_dens1d_labels() and stat_dens1d_filter()

These stats are similar to `stat_dens2d_labels()` and
`stat_dens2d_filter()` but compute the density in a single dimension,
either the *x* or *y* aesthetics.

**Warning!** Which observations are selected by the algorithm used,
based on `stats::density`, depends strongly on the values of parameters
`bw`, `adjust` and `kernel`. You may need to alter the defaults by
passing explicit arguments. Beware that what are good values, may depend
on individual data sets even if they include the same number of
observations. For the selection of observations to work cleanly, the
argument for `n` must large enough to generate a dense grid or the
bandwidth may need to be increased by passing a number \> 1 as argument.
Increasing the bandwidth makes the empirical density function smoother,
and the selection of points less dependent on immediate neighbours.

We use the same data as in the previous sections.

```{r, eval = FALSE}
random_string <- function(len = 6) {
paste(sample(letters, len, replace = TRUE), collapse = "")
}

# Make random data.
set.seed(1001)
d <- tibble::tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A", "B"), c(50, 50)),
  lab = replicate(100, { random_string() })
)
```

The stat `stat_dens1d_filter` *filters* observations, in other words
passes to the geom a subset of the data received as input. The default
value for `geom` is `"point"` and the default `orientation` is `"x"`.

Using defaults except for the color aesthetic, we highlight 1/4 of
observations from lowest density region along the *x* axis of the plot
panel.

```{r}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens1d_filter(keep.fraction = 0.25,
                     colour = "red")
```

We repeat the example above, we highlight 1/4 of observations, but now
from lowest density region along the *y* axis of the plot panel.

```{r, eval=eval_all}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens1d_filter(keep.fraction = 0.25,
                     colour = "red",
                     orientation = "y")
```

In other respects than `orientation` and the parameters passed
internally to `stats::density()` the examples given earlier for
`stat_dens2d_filter()` also apply.

The stat `stat_dens1d_labels` *replaces* the values of the label
(aesthetic) variable in `data` based on density of observations along
the *x* or *y* axis in the plot panel. The replacement is given by the
argument passed to `label.fill`, which can be a character string or a
function accepting a character string as argument and returning also a
character string.

The default value for `geom` is `"text"`. The default value of
`label.fill` is `""` which results in empty labels, while using `NA` as
fill label results in observations being omitted. Using `NA` as
`label.fill` is similar to using `stat_dens2d_filter` as long as the
geom used requires a `label` aesthetic.

Label 1/10 of observations from lowest density regions along *x* in the
plot panel. Normally a repulsive geom is most useful.

```{r, eval=eval_ggrepel}
ggplot(data = d, aes(x, y, label = lab, colour = group)) +
  geom_point() +
  stat_dens1d_labels(geom = "text_repel")
```

Similarly along the *y* axis.

```{r, eval=(eval_all && eval_ggrepel)}
ggplot(data = d, aes(x, y, label = lab, colour = group)) +
  geom_point() +
  stat_dens1d_labels(geom = "text_repel", orientation = "y")
```

In other respects than `orientation` and the parameters passed
internally to `stats::density()` the examples given earlier for
`stat_dens2d_labels()` also apply.

## Simple position functions

Nudging shifts deterministically the coordinates giving an *x* and/or
*y* position and also expands the limits of the corresponding scales to
match. By default in 'ggplot2' geometries and position functions no
nudging is applied.

Function `position_nudge()` from package 'ggplot2' simply applies the
nudge, or *x* and/or *y* shifts based directly on the values passed to
its parameters `x` and `y`. Passing arguments to the `nudge_x` and/or
`nudge_y` parameters of a geometry has the same effect as these values
are passed to `position_nudge()`. Geometries also have a `position`
parameter to which we can pass an expression based on a *position
function* which opens the door to more elaborate approaches to nudging.

A new variation on simple nudge is provided by function
`position_nudge_to()`, which accepts the desired nudged final
coordinates directly instead of as a shift.

We can do better than simply shifting all data to the same extent and
direction or to a fixed position. For example by nudging away from a
focal point, a line or a curve. In `position_nudge_center()` and
`position_nudge_line()` described below, this reference alters only the
direction (angle) along which nudge is applied but not the extent of the
shift. Advanced nudging works very well, but only for some patterns of
observations and may require manual adjustment of positions, repulsion
is more generally applicable but like jitter is aleatory. Combining
nudging and repulsion we can make repulsion more predictable with little
loss of its applicability.

These functions can be used with any geometry but if segments joining
the labels to the points are desired, `ggrepel::geom_text_repel()` or
`ggrepel::geom_label_repel()`, possibly setting `max.iter = 0` if no
repulsion is desired. Several of the functions defined in 'ggpp' even if
not supporting repulsion, can plot connecting segments. Please see the
documentation for these geometries for the details. Drawing of segments
or arrows is made possible by storing in `data` both the nudged and
original *x* and *y* coordinates. This is made possible by coordinated
development of packages 'ggpp' and 'ggrepel' and agreement on a naming
convention for storing the original position.

### position_nudge_keep() and friends

Function `position_nudge_keep()` is like `ggplot2::position_nudge()` but
keeps (stores) the original *x* and *y* coordinates;
`position_nudge_keep()` can be used interchangeably with
`ggplot2::position_nudge()` with other geometries. Although possibly
only occasionally needed, `position_jitter_keep()` ,
`position_dodge_keep()` and `position_stack_keep()` are also made
available for completeness.

```{r}
set.seed(84532)
df <- data.frame(
  x = rnorm(20),
  y = rnorm(20, 2, 2),
  l = paste("label:", letters[1:20])
)

```

When used together with `geom_text_s()` or `geom_label_s()` segments
between a nudged and/or repelled label and the original position (here
indicated by a point) are drawn.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              hjust = "left") +
  expand_limits(x = 2.5)
```

We can supress the drawing of segments.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              hjust = "left", add.segments = FALSE) +
  expand_limits(x = 2.5)
```

With `position_nudge()` used together with `geom_text_repel()` or
`geom_label_repel()` segments connecting a nudged and/or repelled label
and the original position (here indicated by a point) are, by default,
**not** drawn unless they are longer than a minimum.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge(x = 0.3),
                  min.segment.length = 0, 
                  max.iter = 0)
```

With `geom_text_repel()` or `geom_label_repel()` we can force drawing of
segments irespective of their length if we pass `min.segment.length = 0`
in the call.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge_keep(x = 0.3),
                  min.segment.length = 0, max.iter = 0)
```

`position_nudge_keep()` and all other position functions described below
save the original positions in `data` in columns `x_orig` and `y_orig`
and the shifted positions in columns `x` and `y`. Because of this, they
can be used together with any 'ggplot2' geometry, even though these
geometries will ignore the stored original positions.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge(x = 0.3))
```

### position_nudge_to()

Function `position_nudge_to()` nudges to a given position instead of
using the same shift for each observation. Can be used to align labels
for points that are not themselves aligned. By left-justifying the label
text we ensure the alignment of labels differing in width.

```{r}
ggplot(df, aes(x, y, label = ifelse(x < 1, "", l) )) +
  geom_point() +
  geom_text_s(position = position_nudge_to(y = 2.3),
              segment.color = "red",
              arrow = arrow(length = unit(0.015, "npc")),
              angle = 90,
              hjust = "left") +
  expand_limits(x = 3)
```

We here the pass a vector of length one as argument for `y`, but `x` and
`y` also accept vectors of the same length as rows has `data`. In other
words, it allows manual positioning of text and labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point(color = "red", size = 3) +
  geom_text_repel(position = position_nudge_to(y = -2.5), 
                  size = 3,
                  color = "red",
                  angle = 90,
                  hjust = 0,
                  box.padding = 0.1,
                  segment.color = NA,
                  direction = "x") +
  geom_rug(sides = "b", length = unit(0.02, "npc"), color = "red")
```

### position_nudge_center()

Function `position_nudge_center()` can nudge radially away from a focal
point if both `x` and `y` are passed as arguments, or towards opposite
sides of a boundary vertical or horizontal *virtual* line if only one of
`x` or `y` is passed an argument. By default, the "center" is the
centroid computed using `mean()`, but other functions or numeric values
can be passed to override it. When data are sparse, such nudging may be
effective in avoiding label overlaps, and achieving a visually pleasing
positioning.

In all cases nudging shifts the coordinates giving an *x* and/or *y*
position and also expands the limits of the corresponding scales to
include the nudged coordinate values.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3, center_x = 0),
                    min.segment.length = 0, max.iter = 0)
```

By default, split is away or towards the `mean()`. Here we allow
repulsion to separate the labels (compare with previous plot).

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as a constant value.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = 1,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by a function
function, by name.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = median,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by an anonymous
function function. Here que split on the first quartile along *x*.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = function(x) {
                                            quantile(x, 
                                                     probs = 1/4, 
                                                     names = FALSE)
                                          },
                                          direction = "split"),
                  min.segment.length = 0)
```

The labels can be rotated as long as the geometry used supports this.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(angle = 90,
                  position = 
                    position_nudge_center(y = 0.1,
                                          direction = "split"))
```

By requesting nudging along *x* and *y* and setting
`direction = "split"` nudging is applied according to the quadrants
centred on the centroid of the data.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.1,
                                          y = 0.15,
                                          direction = "split"))
```

With `direction = "radial"`, the distance nudged away from the center is
the same for all labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.25,
                                          y = 0.4,
                                          direction = "radial"),
                  min.segment.length = 0)
```

As shown above for `direction = "split"` we can set the coordinates of
the centre by passing constant values as in this example or functions
(possibly different for *x* and *y*). We can also set the justification
the text labels although repulsion usually works best with labels
justified at the centre, which is the default.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.15,
                                          y = 0.3,
                                          center_x = 0,
                                          center_y = 0,
                                          direction = "radial"),
                  min.segment.length = 0,
                  hjust = "outward", vjust = "outward")
```

Nudging along one axis, here *x*, and setting the repulsion `direction`
along the other axis, here *y*, tends to give a pleasent arrangement of
labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.2,
                                          center_x = 0,
                                          direction = "split"),
                  aes(hjust = ifelse(x < 0, 1, 0)),
                  direction = "y",
                  min.segment.length = 0) +
  expand_limits(x = c(-3, 3))
```

### position_nudge_line()

Function `position_nudge_line()` nudges away from a line, which can be a
user supplied straight line as well as a smooth spline or a polynomial
fitted to the observations themselves. The nudging is away and
perpendicular to the local slope of the straight or curved line. It
relies on the same assumptions as linear regression, assuming that *x*
values are not subject to error. This in most cases prevents labels from
overlaping a curve fitted to the data, even if not exactly based on the
same model fit. When observations are sparse, this may be enough to
obtain a nice arrangement of data labels.

```{r}
set.seed(16532)
df <- data.frame(
  x = -10:10,
  y = (-10:10)^2,
  yy = (-10:10)^2 + rnorm(21, 0, 4),
  yyy = (-10:10) + rnorm(21, 0, 4),
  l = letters[1:21]
)
```

The first, simple example shows that `position_nudge_line()` has shifted
the direction of the nudging based on the alignment of the observations
along a line. One could, of course, have in this case passed suitable
values as arguments to *x* and *y* using `position_nudge()` from package
'ggplot2'. However, `position_nudge_line()` will work without change
irrespective of the slope or intercept along which the observations
fall.

```{r}
ggplot(df, aes(x, 2 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 2, linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.5, y = -0.8))
```

With observations with high variation in *y*, a linear model fit may
need to be used. In this case fitted twice, once in `stat_smooth()` and
once in `position_nudge_line()`.

```{r}
ggplot(subset(df, x >= 0), aes(x, yyy)) +
  geom_point() +
  stat_smooth(method = "lm") +
  geom_text(aes(label = l),
            vjust = "center", hjust = "center",
            position = position_nudge_line(x = 0, y = 1.2,
                                           method = "lm",
                                           direction = "split"))
```

With lower variation in *y*, we can pass to `line_nudge` a multiplier to
keep labels outside of the confidence band.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            position = position_nudge_line(method = "lm",
                                           x = 3, y = 3, 
                                           line_nudge = 2.5,
                                           direction = "split"))
```

If we want the nudging based on an arbitrary straight line not computed
from `data`, we can pass the intercept and slope in a numeric vector of
length two as an argument to parameter `abline`.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  geom_text(aes(label = l),
            position = position_nudge_line(abline = c(0, 1),
                                           x = 3, y = 3, 
                                           direction = "split"))
```

With observations that follow exactly a simple curve the defaults work
well to automate the nudging of individual data labels away from the
implicit curve. Positive values as arguments to `x` and `y` correspond
to above and inside the curve. One could, of course, pass also in this
case suitable values as arguments to *x* and *y* using
`position_nudge()` from package 'ggplot2', but these arguments would
need to be vectors with different *nudge* values for each observation.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = 0.6, y = 6))
```

Negative values passed as arguments to `x` and `y` correspond to labels
below and outside the curve.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.6, y = -6))
```

When the observations include random variation along *y*, it is
important that the smoother used for the line added to a plot and that
passed to `position_nudge_line()` are similar. By default
`stat_smooth()` uses `"loess"` and `position_nudge_line()` with method
`"spline"`, `smooth.sline()`, which are a good enough match.

```{r}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth() +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(x = 0.6, 
                                           y = 6,
                                           direction = "split"))
```

We can use other geometries.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth() +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
            position = position_nudge_line(x = 0.6, 
                                           y = 6,
                                           direction = "split"),
            min.segment.length = 0)
```

When fitting a polynomial, `"lm"` should be the argument passed to
`method` and a model formula preferably based on `poly()`, setting
`raw = TRUE`, as argument to `formula`. *Currently no other methods are
implemented in* `position_nudge_line()`.

```{r}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(method = "lm",
                                           x = 0.5, 
                                           y = 5,
                                           formula = y ~ poly(x, 2, raw = TRUE),
                                           direction = "split"))
```

Compared to pure repulsion adding nudging helps "direct" repulsion in
the desired direction. Compare the two plots bellow using no nudging
with the two above.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text_repel(aes(y = yy, label = l),
                  box.padding = 0.5,
                  min.segment.length = Inf)
```

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth() +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
                  box.padding = 0.5,
                  min.segment.length = 0)
```

Grouping is supported.

```{r}
df <- data.frame(x = rep(1:10, 2),
                 y = c(1:10, 10:1),
                 group = rep(c("a", "b"), c(10, 10)),
                 l = "+")
```

```{r}
ggplot(df, aes(x, y, label = l, color = group)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(position = position_nudge_line(x = 0.25, y = 1)) +
  geom_text(position = position_nudge_line(x = -0.25, y = -1)) +
  coord_equal(ratio = 0.5)
```

One needs to ensure that grouping is in effect in the geoms with
nudging.

```{r}
ggplot(df, aes(x, y, label = l, color = group, group = group)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(color = "red",
            position = position_nudge_line(x = 1, y = 1)) +
  geom_text(color = "blue",
            position = position_nudge_line(x = -1, y = -1)) +
  coord_equal()
```

Facets are also supported.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(position = position_nudge_line(x = 1, y = 1),
            color = "red") +
  geom_text(position = position_nudge_line(x = -1, y = -1),
            color = "blue") +
  facet_wrap(~group) +
  coord_equal(ratio = 1)
```

## Combined position functions

Nudging text or labels in a dodged or stacked bar or column plot has not
been easy in 'ggplot2'. This is a rather frequent situation, although
usually there are workarounds. Non-the-less this imposed some
limitations on the design of plots. The examples below show the use of
*position* that implement the functionality of more than one of the
position functions available in 'ggplot2'. ***Examples will be added
here for the next release. Meanwhile, please, read the help page for
each function. These pages contain examples.***

## Appendix A: More on density filtering

We define a function to simplify the generation of random data sets
based on different probability distributions.

```{r}
make_data_tbl <- function(nrow = 100, rfun = rnorm, ...) {
  if (nrow %% 2) {
    nrow <- nrow + 1
  }
  
  set.seed(1001)
  
  tibble::tibble(
    x = rfun(nrow, ...),
    y = rfun(nrow, ...),
    group = rep(c("A", "B"), c(nrow / 2, nrow / 2))
  )
}
```

In all the examples in this vignette we use colors to demonstrate which
data points are selected, but any other suitable aesthetic and discrete
scale can be used instead. With `keep.sparse = FALSE` we keep 1/3 of the
observations in the denser region of the plot. Although here we first
plot all data points and later overplot the selected ones this is not
necessary.

```{r, eval=eval_all}
ggplot(data = make_data_tbl(300), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(colour = "red", 
                     keep.sparse = FALSE, 
                     keep.fraction = 1/3)
```

Here we highlight the observations split in three group equal groups,
each of a different density of observations.

```{r, eval=eval_all}
ggplot(data = make_data_tbl(300), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(colour = "red", 
                     keep.sparse = FALSE, 
                     keep.fraction = 1/3)+
  stat_dens2d_filter(colour = "blue", 
                     keep.fraction = 1/3)
```

The algorithm seems to work well also with other distributions, in this
example the uniform distribution.

```{r, eval=eval_all}
ggplot(data = make_data_tbl(300, rfun = runif), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(colour = "red", keep.fraction = 1/2)
```

One example with the gamma distribution, which is asymmetric.

```{r, eval=eval_all}
ggplot(data = make_data_tbl(300, rfun = rgamma, shape = 2), 
       aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(colour = "red", keep.fraction = 1/3)
```

## Appendix B: try_tibble()

### Time series

Several different formats for storing time series data are used in R.
Here we use in the examples objects of class `ts` but several other
classes are supported as `try.xts()` is used internally. The first
example is a quarterly series.

```{r}
class(austres)
austres.df <- try_tibble(austres)
class(austres.df)
head(austres.df, 4)
```

The next chunk demonstrates that numeric times are expressed as decimal
years in the returned data frame.

```{r}
austres.df <- try_tibble(austres, as.numeric = TRUE)
head(austres.df, 4)
```

This second example is for a series of yearly values.

```{r, eval=eval_all}
class(lynx)
lynx.df <- try_tibble(lynx)
class(lynx.df)
head(lynx.df, 3)
```

Above there is a small rounding error of 1 s for these old dates. We can
correct this by rounding to year.

```{r}
lynx.df <- try_tibble(lynx, "year")
head(lynx.df, 3)
```

In addition we can convert the POSIXct values into numeric values in
calendar years plus a decimal fraction.

```{r, eval=eval_all}
lynx_n.df <- try_tibble(lynx, "year", as.numeric = TRUE)
lapply(lynx_n.df, "class")
head(lynx_n.df, 3)
```

### Other classes

`try_tibble()` attempts to handle gracefully objects that are not time
series.

```{r, eval=eval_all}
try_tibble(1:5)
```

```{r, eval=eval_all}
try_tibble(letters[1:5])
```

```{r, eval=eval_all}
try_tibble(factor(letters[1:5]))
```

```{r, eval=eval_all}
try_tibble(list(x = rep(1,5), y = 1:5))
```

```{r, eval=eval_all}
try_tibble(data.frame(x = rep(1,5), y = 1:5))
```

```{r, eval=eval_all}
try_tibble(matrix(1:10, ncol = 2))
```
