---
title: "Combining repulsion and nudging"
subtitle: "Packages 'ggrepel' and 'ggpp' working as a team"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Combining repulsion and nudging}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
```

## Introduction

The very popular R package 'ggrepel' does a great job at avoiding overlaps among
data labels and between them and observations plotted as points. A difficulty
that stems from the use of an algorithm based on random displacements is that
the final location of the data labels can become more disordered than necessary.
In addition when including smooth regression lines the data labels may partly
occlude the fitted line and/or the confidence band.

Package 'ggpp' defines new position functions that save the starting position
like `position_nudge_repel()` does but come in multiple flavors. Their use
together with repulsive geometries from 'ggrepel' makes it possible to give to
the data labels an initial "push" in a non-random direction. This helps a lot,
much more than what I expect initially, in obtaining more orderly
displacement of the data labels away from a cloud of observations or line.

Another problem sometimes encountered when using position functions is that
combinations of pairs of displacements would be required. 'ggpp' does define
such new position functions which can also be used together with repulsive
geometries from package 'ggrepel'.

Because of the naming convention used, the new position functions remain fully
compatible with all geometries that have a formal parameter `position`. However,
most examples below use geometries from packages 'ggrepel' or 'ggpp' to
create a plot layer containing data labels.

## Preliminaries

As we will use text and labels on the plotting area we change the
default theme to an uncluttered one.

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggpp)
library(ggrepel)

old_theme <- theme_set(theme_bw())
```

## Nudge position functions

Nudging shifts deterministically the coordinates giving an _x_ and/or _y_
position and also expands the limits of the corresponding scales to match. By
default in 'ggplot2' geometries and position functions no nudging is applied.

Function `position_nudge()` from package 'ggplot2' simply applies the
nudge, or *x* and/or *y* shifts based directly on the values passed to its
parameters `x` and `y`. Passing arguments to the `nudge_x` and/or `nudge_y`
parameters of a geometry has the same effect as these values are passed to
`position_nudge()`. Geometries also have a `position` parameter to which
we can pass an expression based on a _position function_ which opens the
door to more elaborate approaches to nudging.

A new variation on simple nudge is provided by function `position_nudge_to()`,
which accepts the desired nudged coordinates directly.

We can do better than simply shifting all data to the same extent and direction
or to a fixed position. For example by nudging away from a focal point, a line
or a curve.  In `position_nudge_center()` and `position_nudge_liner()` described
below, this reference alters only the direction (angle) along which nudge is
applied but not the extent of the shift. Advanced nudging works very well, but
only for some patterns of observations and may require manual adjustment of
positions, repulsion is more generally applicable but like jittering is
aleatory. Combining nudging and repulsion we can make repulsion more predictable
with little loss of its applicability.

These functions can be used with any geometry but if segments joining the labels
to the points are desired, `ggrepel::geom_text_repel()` or
`ggrepel::geom_label_repel()` should be used, possibly setting `max.iter = 0` if
no repulsion is desired. Plotting of short segments can be forced with
`min.segment.length = 0`. Please see the documentation for these geometries for
the details. Drawing of segments or arrows is made possible by storing in `data`
both the nudged and original _x_ and _y_ coordinates. This is made possible by
coordinated development of packages 'ggpp' and 'ggrepel' and a naming
convention for storing the original position. 

### position_nudge_keep()

Function `position_nudge_keep()` is like `ggplot2::position_nudge()` but keeps
(stores) the original _x_ and _y_ coordinates. It is similar to function
`position_nudge_repel()` but uses a different naming convention for the
coordinates. Both work with `geom_text_repel()` or `geom_label_repel()` from
package 'ggrepel' (> 0.9.1), but only `position_nudge_keep()` can be used
interchangeably with `ggplot2::position_nudge()` with other geometries.

```{r}
set.seed(84532)
df <- data.frame(
  x = rnorm(20),
  y = rnorm(20, 2, 2),
  l = paste("label:", letters[1:20])
)

```

With `position_nudge_keep()` from 'ggpp' used together with `geom_text_repel()` or 
`geom_label_repel()` segments between a nudged and/or repelled label and the
original position (here indicated by a point) are drawn. As shown here, passing
`max.iter = 0` disables repulsion.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge_keep(x = 0.3),
                  min.segment.length = 0, max.iter = 0)
```

With `position_nudge()` from 'ggplot2' used together with `geom_text_repel()` or 
`geom_label_repel()` segments connecting a nudged and/or repelled label and
the original position (here indicated by a point) are **not** drawn.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge(x = 0.3),
                  min.segment.length = 0, 
                  max.iter = 0)
```

`position_nudge_keep()` and all other position functions described below can
be used with all 'ggplot2' geometries but the original position will be ignored
and no connecting segment drawn unless the geometry has been designed to work
together with them. Currently, `geom_text_repel()` and `geom_label_repel()` from
'ggrepel' and `geom_text_s()`, `geom_label_s()` and `geom_point_s()` from package
'ggpp' draw connecting segments.

`geom_text_s()`, `geom_label_s()` and `geom_point_s()` are different in some respects from `geom_text_repel()` and `geom_label_repel()`, adding a few features and lacking several of those available in `geom_text_repel()` and `geom_label_repel()`, in which as shown above repulsion can be disabled if desired. `geom_text_s()`, `geom_label_s()` and `geom_point_s()` do not support repulsion and differ in that aesthetic mappings can be selectively applied to the different components of the label and/or to segments. 

In addition, `geom_text_s()` and `geom_label_s()` use a different approach to justification than `geom_text_repel()`, these geometries justify the text or label to the nearest edge to the original position, while `geom_text_repel()` by default centres the text label on the nudged or displaced coordinates.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              min.segment.length = 0) +
  expand_limits(x = 2.3)
```

`geom_text_repel()` trims the segment to the edge of the text plus the padding. In contrast, `geom_text_s()` uses justification to avoid the overlap and only the default justification `"position"` and one of the edges, "left" in this case, are currently usable.
 
```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              min.segment.length = 0,
              hjust = "left") +
  expand_limits(x = 2.3)
```
A usually more problematic example is the labeling of loadings in PCA and similar biplots.

```{r}
## Example data frame where each species' principal components have been computed.
df1 <- data.frame(
  Species = paste("Species",1:5),
  PC1     = c(-4, -3.5, 1, 2, 3),
  PC2     = c(-1, -1, 0, -0.5, 0.7)) 

ggplot(df1, aes(x=PC1, y = PC2, label = Species, colour = Species)) +
  geom_hline(aes(yintercept = 0), size=.2) +
  geom_vline(aes(xintercept = 0), size=.2) +
  geom_segment(aes(x=0, y=0, xend=PC1, yend=PC2), 
               arrow = arrow(length = unit(0.1, "inches"))) +
  geom_label_repel(position = position_nudge_center(x = 0.2, y = 0.01,
                                                    center_x = 0, center_y = 0),
                   label.size = NA,
                   label.padding = 0.1,
                   fill = rgb(red = 1, green = 1, blue = 1, alpha = 0.75)) +
  xlim(-5, 5) +
  ylim(-2, 2) +
  # Stadard settings for displaying biplots
  coord_fixed() +
  theme(legend.position = "none")
```
### position_nudge_to()

Function `position_nudge_to()` nudges to a given position instead of using
the same shift for each observation. Can be used to align labels for points 
that are not themselves aligned.

```{r}
ggplot(df, aes(x, y, label = ifelse(x < 0.5, "", l) )) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_to(x = 2.3),
                  min.segment.length = 0,
                  segment.color = "red",
                  arrow = arrow(length = unit(0.015, "npc")),
                  direction = "y") +
  expand_limits(x = 3)
```
By providing to values for nudging with opposite sign, we can add labels alternating between sides. We use here `geom_text_s()` but other geometries could have been used instead.

```{r}
size_from_area <- function(x) {sqrt(max(0, x) / pi)}

df2 <- data.frame(b = exp(seq(2, 4, length.out = 10)))

ggplot(df2, aes(1, b, size = b)) + 
  geom_text_s(aes(label = round(b,2)),
              position = position_nudge_to(x = c(1.1, 0.9)),
              box.padding = 0) +
  geom_point() +
  scale_size_area() +
  xlim(0, 2) +
  theme(legend.position = "none")
```

It is useful when labeling curves than end at different positions along the _x_ axis.

```{r}
keep <- c("Israel", "United States", "European Union", "China", "South Africa", "Qatar",
          "Argentina", "Chile", "Brazil", "Ukraine", "Indonesia", "Bangladesh")

data <- read.csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv")
data$date <- ymd(data$date)

data %>%
  filter(location %in% keep) %>%
  select(location, date, total_vaccinations_per_hundred) %>%
  arrange(location, date) %>%
  filter(!is.na(total_vaccinations_per_hundred)) %>%
  mutate(location = factor(location),
         location = reorder(location, total_vaccinations_per_hundred)) %>%
  group_by(location) %>% # max(date) depends on the location!
  mutate(label = if_else(date == max(date), 
                         as.character(location), 
                         "")) -> owid

ggplot(owid,
       aes(x = date, 
           y = total_vaccinations_per_hundred,
           color = location)) +
  geom_line() +
  geom_text_repel(aes(label = label),
                  size = 3,
                  position = position_nudge_to(x = max(owid$date) + days(30)),
                  segment.color = 'grey',
                  point.size = 0,
                  box.padding = 0.1,
                  point.padding = 0.1,
                  hjust = "left",
                  direction="y") + 
  scale_x_date(expand = expansion(mult = c(0.05, 0.2))) +
  labs(title = "Cumulative COVID-19 vaccination doses administered per 100 people",
       y = "",
       x = "Date (year-month)") +
  theme_bw() +
  theme(legend.position = "none")
```

We here the pass a vector of length one as argument for `y`, but both `x` and `y`
also accept longer vectors. In other words, it allows manual positioning of text 
and labels.

In the next example we decrease the forces used for repulsion and the padding so
that the labels remain close together. In this way, we can label the 
observations on the rug of a combined point and rug plot.
 
```{r}
ggplot(df, aes(x, y, label = round(x, 2))) +
  geom_point(color = "red", size = 3) +
  geom_text_repel(position = position_nudge_to(y = -2.7), 
            size = 3,
            color = "red",
            angle = 90,
            hjust = 0,
            box.padding = 0.05,
            min.segment.length = Inf,
            direction = "x",
            force = 0.1,
            force_pull = 0.1) +
  geom_rug(sides = "b", length = unit(0.02, "npc"), color = "red")
```

### position_nudge_center()

Function `position_nudge_center()` can nudge radially away from a focal point if
both `x` and `y` are passed as arguments, or towards opposite sides of a
boundary vertical or horizontal _virtual_ line if only one of `x` or `y` is
passed an argument. By default, the "center" is the centroid computed using
`mean()`, but other functions or numeric values can be passed to override it.
When data are sparse, such nudging may be effective in avoiding label overlaps,
and achieving a visually pleasing positioning.

In all cases nudging shifts the coordinates giving an _x_ and/or _y_ position
and also expands the limits of the corresponding scales to include the nudged
coordinate values.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3, center_x = 0),
                    min.segment.length = 0, max.iter = 0)
```

By default, split is away or towards the `mean()`. Here we allow repulsion to 
separate the labels (compare with previous plot).

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as a constant value.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = 1,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by a function function,
by name.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = median,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by an anonymous function 
function. Here que split on the first quartile along _x_.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = function(x) {
                                            quantile(x, 
                                                     probs = 1/4, 
                                                     names = FALSE)
                                          },
                                          direction = "split"),
                  min.segment.length = 0)
```

The labels can be rotated as long as the geometry used supports this.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(angle = 90,
                  position = 
                    position_nudge_center(y = 0.1,
                                          direction = "split"))
```

By requesting nudging along _x_ and _y_ and setting `direction = "split"` nudging is applied according to the quadrants centred on the centroid of the data.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.1,
                                          y = 0.15,
                                          direction = "split"))
```

With `direction = "radial"`, the distance nudged away from the center is 
the same for all labels.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.25,
                                          y = 0.4,
                                          direction = "radial"),
                  min.segment.length = 0)
```
As shown above for `direction = "split"` we can set the coordinates of the 
center also with `direction = "radial"`. In the case of `geom_text_s()` the
default justification is `"position"`.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = 
                    position_nudge_center(x = 0.125,
                                          y = 0.2,
                                          center_x = -0.5,
                                          direction = "radial"),
                  min.segment.length = 0) +
  expand_limits(x = c(-2.7, +2.3))
```

We can also set the justification the text labels
although repulsion usually works best with labels justified at the centre,
which is the default in `geom_text_repel()`.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.125,
                                          y = 0.25,
                                          center_x = 0,
                                          center_y = 0,
                                          direction = "radial"),
                  min.segment.length = 0,
                  hjust = "outward", vjust = "outward") +
  expand_limits(x = c(-2.7, +2.3))
```

Nudging along one axis, here _x_, and setting the repulsion `direction` along
the other axis, here _y_, tends to give a pleasant arrangement of labels.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.2,
                                          center_x = 0,
                                          direction = "split"),
                  aes(hjust = ifelse(x < 0, 1, 0)),
                  direction = "y",
                  min.segment.length = 0) +
  expand_limits(x = c(-3, 3))
```

### position_nudge_line()

Function `position_nudge_line()` nudges away from a line, which can be a
user supplied straight line as well as a smooth spline or a polynomial
fitted to the observations themselves. The nudging is away and
perpendicular to the local slope of the straight or curved line. It
relies on the same assumptions as linear regression, assuming that _x_
values are not subject to error. This in most cases prevents labels from
overlaping a curve fitted to the data, even if not exactly based on the
same model fit. When observations are sparse, this may be enough to
obtain a nice arrangement of data labels.

```{r}
set.seed(16532)
df <- data.frame(
  x = -10:10,
  y = (-10:10)^2,
  yy = (-10:10)^2 + rnorm(21, 0, 4),
  yyy = (-10:10) + rnorm(21, 0, 4),
  l = letters[1:21]
)
```

The first, simple example shows that `position_nudge_line()` has shifted the
direction of the nudging based on the alignment of the observations along a
line. One could, of course, have in this case passed suitable values as
arguments to *x* and *y* using `position_nudge()` from package 'ggplot2'.
However, `position_nudge_line()` will work without change irrespective of the
slope or intercept along which the observations fall.

```{r}
ggplot(df, aes(x, 2 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 2, linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.5, y = -0.8))
```

With observations with high variation in *y*, a linear model fit may
need to be used. In this case fitted twice, once in `stat_smooth()` and once in
`position_nudge_line()`.

```{r}
ggplot(subset(df, x >= 0), aes(x, yyy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            vjust = "center", hjust = "center",
            position = position_nudge_line(x = 0, y = 1.2,
                                           method = "lm",
                                           direction = "split"))
```

With lower variation in *y*, we can pass to `line_nudge` a multiplier to
keep labels outside of the confidence band.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            position = position_nudge_line(method = "lm",
                                           x = 3, y = 3, 
                                           line_nudge = 2.5,
                                           direction = "split"))
```

If we want the nudging based on an arbitrary straight line not computed
from `data`, we can pass the intercept and slope in a numeric vector of
length two as an argument to parameter `abline`.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  geom_text(aes(label = l),
            position = position_nudge_line(abline = c(0, 1),
                                           x = 3, y = 3, 
                                           direction = "split"))
```

With observations that follow exactly a simple curve the defaults work well to
automate the nudging of individual data labels away from the implicit curve.
Positive values as arguments to `x` and `y` correspond to above and inside the
curve. One could, of course, pass also in this case suitable values as arguments
to *x* and *y* using `position_nudge()` from package 'ggplot2', but these
arguments would need to be vectors with different *nudge* values for each
observation.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = 0.6, y = 6))
```

Negative values passed as arguments to `x` and `y` correspond to labels
below and outside the curve.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.6, y = -6))
```

When the observations include random variation along *y*, it is
important that the smoother used for the line added to a plot and that
passed to `position_nudge_line()` are similar. By default
`stat_smooth()` uses `"loess"` and `position_nudge_line()` with method
`"spline"`, `smooth.sline()`, which are a good enough match.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(x = 0.6, 
                                           y = 6,
                                           direction = "split"))
```

We can use other geometries, or rather we need to use a repulsive geometry when the label text is long or the labels are crowed near the line.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
                   position = position_nudge_line(x = 0.6, 
                                                  y = 8,
                                                  direction = "split"),
                   box.padding = 0.2,
                   min.segment.length = 0)
```
We can see by comparing the plot above with that below, that combining nudging away from a line with repulsion results in a more pleasant positioning of the data labels.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
                  box.padding = 0.5,
                  min.segment.length = 0)
``` 

When fitting a polynomial, `"lm"` should be the argument passed to
`method` and a model formula preferably based on `poly()`, setting `raw = TRUE`,
as argument to `formula`.

*Currently no other methods are implemented in* `position_nudge_line()`.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(method = "lm",
                                           formula = y ~ poly(x, 2, raw = TRUE),
                                           x = 0.5, 
                                           y = 5,
                                           direction = "split"))
```

Compared to pure repulsion adding nudging helps "direct" repulsion in the desired direction. Compare the plot below using no nudging with the one above.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text_repel(aes(y = yy, label = l),
                  box.padding = 0.5,
                  min.segment.length = Inf)
```

## Combined position functions

Using `position_stacknudge()` together `geom_label_repel()` makes it possible
to use repulsion for labeling sections of stacked column plots.

```{r}
df <- tibble::tribble(
  ~y, ~x, ~grp,
  "a", 1,  "some long name",
  "a", 2,  "other name",
  "b", 1,  "some name",
  "b", 3,  "another name",
  "b", -1, "some long name"
)

ggplot(data = df, aes(x, y, group = grp)) +
  geom_col(aes(fill = grp), width=0.5) +
  geom_vline(xintercept = 0) +
  geom_label_repel(aes(label = grp),
                   position = position_stacknudge(vjust = 0.5, y = 0.4),
                   label.size = NA)
```


## Acknowledgements

I warmly thank Kamil Slowikowski for agreeing to make changes in 'ggrepel' that make the use of 'ggrepel' together with 'ggpp' possible and smooth. This document shows some use examples, but surely new ones will be found by users of R and 'ggplot2'. 

