---
title: "Combining repulsion and nudging"
subtitle: "Packages 'ggrepel' and 'ggpp' working as a team"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Combining repulsion and nudging}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
```

## Introduction

The very popular R package 'ggrepel' does a great job at avoiding
overlaps among data labels and between them and observations plotted as
points. A difficulty that stems from the use of an algorithm based on
random displacements is that the final location of the data labels can
become more disordered than necessary. In addition when including smooth
regression lines the data labels may partly occlude the fitted line
and/or the confidence band.

Package 'ggpp' defines new position functions that save the starting
position like `position_nudge_repel()` does but come in multiple
flavors. Their use together with repulsive geometries from 'ggrepel'
makes it possible to give to the data labels an initial "push" in a
non-random direction. This helps a lot, much more than what I expect
initially, in obtaining a more orderly displacement by repulsion of the
data labels away from a cloud of observations or a line.

Another problem sometimes encountered when using position functions is
that combinations of pairs of displacements would be required. 'ggpp'
does define such new position functions which can also be used together
with the repulsive geometries from package 'ggrepel'.

Because of the naming convention used, the new position functions remain
fully compatible with all geometries that have a formal parameter
`position`. However, most examples below use geometries from packages
'ggrepel' or 'ggpp' to create a plot layer containing data labels.

## Preliminaries

As we will use text and labels on the plotting area we change the
default theme to an uncluttered one.

```{r, message = FALSE}
# Are the packages used in examples installed?
eval_ggrepel <- requireNamespace("ggrepel", quietly = TRUE)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggpp)
if (eval_ggrepel) library(ggrepel)

old_theme <- theme_set(theme_bw())
```

## Nudge position functions

Nudging shifts deterministically the *x* and/or *y* coordinates of an
observation. This takes place early enough for the limits of the
corresponding scales be set based on the displaced positions. In
'ggplot2', position functions and consequently also geometries by
default apply no nudging.

Function `position_nudge()` from package 'ggplot2' applies the nudge, to
*x* and/or *y* data coordinates based directly on the values passed to
its parameters `x` and `y`. Passing arguments to the `nudge_x` and/or
`nudge_y` parameters of a geometry has the same effect, as these values
are passed to `position_nudge()` within the geometry's code. Geometries
also have a `position` parameter to which we can pass an expression
based on a *position function* which opens the door to more elaborate
approaches to nudging, as well as allowing other changes in coordinates
such as stacking.

A new variation on simple nudge is provided by function
`position_nudge_to()`, which accepts the desired nudged coordinates
directly, instead of as a displacement away from the initial position.
This allows to push data labels away from observations into a row or
column.

We can do better than simply shifting all data to the same extent and
direction or to a fixed position. For example by nudging away from a
focal point, a line or a curve. In `position_nudge_center()` and
`position_nudge_liner()` described below, this reference alters only the
direction (angle) along which nudge is applied but not the extent of the
shift. Advanced nudging works very well, but only for some patterns of
observations and may require manual adjustment of positions, repulsion
is more generally applicable but like jittering is aleatory. Combining
nudging and repulsion we can make repulsion more predictable with little
loss of its applicability.

These functions can be used with any geometry but if segments joining
the labels to the points are desired, `ggrepel::geom_text_repel()` or
`ggrepel::geom_label_repel()` should be used when repulsion is desired.
Setting `max.iter = 0` in these functions disables repulsion.
Alternatively, several geometries from 'ggpp' implement the drawing of
connecting segments. Please see the documentation for these geometries
for the details.

Drawing of segments or arrows is made possible by storing in `data` both
the nudged and original *x* and *y* coordinates. This was made possible
by coordinated development of packages 'ggpp' and 'ggrepel' and
agreement on a naming convention for storing the original position.

### position_nudge_keep()

Function `position_nudge_keep()` is like `ggplot2::position_nudge()` but
keeps (stores) the original *x* and *y* coordinates. It is similar to
function `position_nudge_repel()` but uses a different naming convention
for the coordinates. Both work with `geom_text_repel()` or
`geom_label_repel()` from package 'ggrepel' (\> 0.9.1), but only
`position_nudge_keep()` can be used interchangeably with
`ggplot2::position_nudge()` with other geometries.

```{r}
set.seed(84532)
df <- data.frame(
  x = rnorm(20),
  y = rnorm(20, 2, 2),
  l = paste("label:", letters[1:20])
)

```

With `position_nudge_keep()` from 'ggpp' used together with
`geom_text_repel()` or `geom_label_repel()` segments between a nudged
and/or repelled label and the original position (here indicated by a
point) are drawn. As shown here, passing `max.iter = 0` disables
repulsion.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge_keep(x = 0.3),
                  min.segment.length = 0, max.iter = 0)
```

With `position_nudge()` from 'ggplot2' used together with
`geom_text_repel()` or `geom_label_repel()` segments connecting a nudged
and/or repelled label and the original position (here indicated by a
point) are **not** drawn.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = position_nudge(x = 0.3),
                  min.segment.length = 0, 
                  max.iter = 0)
```

`position_nudge_keep()` and all other position functions described below
can be used with all 'ggplot2' geometries but the original position will
be ignored and no connecting segment drawn unless the geometry has been
designed to work together with them. Currently, `geom_text_repel()` and
`geom_label_repel()` from 'ggrepel' and `geom_text_s()`,
`geom_label_s()` and `geom_point_s()` from package 'ggpp' draw
connecting segments.

`geom_text_s()`, `geom_label_s()` and `geom_point_s()` are different in
some respects from `geom_text_repel()` and `geom_label_repel()`, adding
a few features and lacking several of those available in
`geom_text_repel()` and `geom_label_repel()`, in which as shown above
repulsion can be disabled if desired. `geom_text_s()`, `geom_label_s()`
and `geom_point_s()` do not support repulsion and differ in that
aesthetic mappings can be selectively applied to the different
components of the label and/or to segments while supporting no new
aesthetics. While `geom_text_repel()` and `geom_label_repel()` support
curved connecting segments and arrows, the geometries from 'ggpp'
support only straight segments and arrows.

In addition, `geom_text_s()` and `geom_label_s()` use a different
approach to justification of the displaced data labels than
`geom_text_repel()`. These geometries from 'ggpp' by default justify the
text or label to the nearest edge to the original position (away from
it). A justification approach named `"position"` in 'ggpp' and not available
in geometries defined in other packages.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              min.segment.length = 0) +
  expand_limits(x = 2.3)
```

`geom_text_repel()` draws the segment by default from the centre of the
text and trims it to the edge of the text plus the padding. In contrast,
`geom_text_s()` uses justification to avoid the overlap and only the
default justification `"position"` and one of the edges, "left" in this
case, are currently usable as the untrimmed segment otherwise overplots
the text.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = position_nudge_keep(x = 0.1),
              min.segment.length = 0,
              hjust = "left") +
  expand_limits(x = 2.3)
```

---
Each approach has advantages and disadvantages. The main difference is that 
with `geom_text_s()` and `geom_label_s()` shorter segments are needed than with
`geom_text_repel()` and `geom_label_repel()` when using their default 
justification approach.
---

A usually more problematic example is the labeling of loadings in PCA
and similar biplots.

```{r, eval=eval_ggrepel}
## Example data frame where each species' principal components have been computed.
df1 <- data.frame(
  Species = paste("Species",1:5),
  PC1     = c(-4, -3.5, 1, 2, 3),
  PC2     = c(-1, -1, 0, -0.5, 0.7)) 

ggplot(df1, aes(x=PC1, y = PC2, label = Species, colour = Species)) +
  geom_hline(aes(yintercept = 0), size=.2) +
  geom_vline(aes(xintercept = 0), size=.2) +
  geom_segment(aes(x=0, y=0, xend=PC1, yend=PC2), 
               arrow = arrow(length = unit(0.1, "inches"))) +
  geom_label_repel(position = position_nudge_center(x = 0.2, y = 0.01,
                                                    center_x = 0, center_y = 0),
                   label.size = NA,
                   label.padding = 0.1,
                   fill = rgb(red = 1, green = 1, blue = 1, alpha = 0.75)) +
  xlim(-5, 5) +
  ylim(-2, 2) +
  # Stadard settings for displaying biplots
  coord_fixed() +
  theme(legend.position = "none")
```

### position_nudge_to()

Function `position_nudge_to()` nudges to a given position instead of
using the same shift for each observation. Can be used to align labels
for points that are not themselves aligned.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = ifelse(x < 0.5, "", l) )) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_to(x = 2.3),
                  min.segment.length = 0,
                  segment.color = "red",
                  arrow = arrow(length = unit(0.015, "npc")),
                  direction = "y") +
  expand_limits(x = 3)
```

By providing to values for nudging with opposite sign, we can add labels
alternating between sides. We use here `geom_text_s()` but other
geometries could have been used as well.

```{r}
size_from_area <- function(x) {sqrt(max(0, x) / pi)}

df2 <- data.frame(b = exp(seq(2, 4, length.out = 10)))

ggplot(df2, aes(1, b, size = b)) + 
  geom_text_s(aes(label = round(b,2)),
              position = position_nudge_to(x = c(1.1, 0.9)),
              box.padding = 0.5) +
  geom_point() +
  scale_size_area() +
  xlim(0, 2) +
  theme(legend.position = "none")
```

It is also useful when labeling curves than end at different positions along
the *x* axis.

```{r, eval=eval_ggrepel}
keep <- c("Israel", "United States", "European Union", "China", "South Africa", "Qatar",
          "Argentina", "Chile", "Brazil", "Ukraine", "Indonesia", "Bangladesh")

data <- read.csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv")
data$date <- ymd(data$date)

data %>%
  filter(location %in% keep) %>%
  select(location, date, total_vaccinations_per_hundred) %>%
  arrange(location, date) %>%
  filter(!is.na(total_vaccinations_per_hundred)) %>%
  mutate(location = factor(location),
         location = reorder(location, total_vaccinations_per_hundred)) %>%
  group_by(location) %>% # max(date) depends on the location!
  mutate(label = if_else(date == max(date), 
                         as.character(location), 
                         "")) -> owid

ggplot(owid,
       aes(x = date, 
           y = total_vaccinations_per_hundred,
           color = location)) +
  geom_line() +
  geom_text_repel(aes(label = label),
                  size = 3,
                  position = position_nudge_to(x = max(owid$date) + days(30)),
                  segment.color = 'grey',
                  point.size = 0,
                  box.padding = 0.1,
                  point.padding = 0.1,
                  hjust = "left",
                  direction = "y") + 
  scale_x_date(expand = expansion(mult = c(0.05, 0.2))) +
  labs(title = "Cumulative COVID-19 vaccination doses administered per 100 people",
       y = "",
       x = "Date (year-month)") +
  theme_bw() +
  theme(legend.position = "none")
```

In the call to `position_nudge_to()` we passed a vector of length one as
argument for `y`, but both `x` and `y` also accept longer vectors. In other
words, this position functions makes it posible manual positioning of text and
labels.

In the next example we decrease the forces used for repulsion and the
padding so that the labels remain close together. In this way, we can
label the observations on the rug of a combined point and rug plot.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = round(x, 2))) +
  geom_point(color = "red", size = 3) +
  geom_text_repel(position = position_nudge_to(y = -2.7), 
            size = 3,
            color = "red",
            angle = 90,
            hjust = 0,
            box.padding = 0.05,
            min.segment.length = Inf,
            direction = "x",
            force = 0.1,
            force_pull = 0.1) +
  geom_rug(sides = "b", length = unit(0.02, "npc"), color = "red")
```

### position_nudge_center()

Function `position_nudge_center()` can nudge radially away from a focal point if
both `x` and `y` are passed as arguments, or towards opposite sides of a
vertical or horizontal *virtual* boundary line if only one of `x` or `y` is
passed an argument. By default, the "center" is the centroid computed using
`mean()`, but other functions or numeric values can be passed to override it.
When data are sparse, such nudging may be effective in avoiding label overlaps,
and in achieving a visually pleasing positioning.

In all cases nudging shifts the coordinates giving a new *x* and/or *y* position
that expands the limits of the corresponding scales to include the nudged
coordinate values, but not necessarily the whole of justified text or labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3, center_x = 0),
                    min.segment.length = 0, max.iter = 0)
```

By default, split is away or towards the `mean()`. Here we allow
repulsion to separate the labels (compare with the previous plot).

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as a constant value.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = 1,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by a function
function, by name.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = median,
                                          direction = "split"),
                  min.segment.length = 0)
```

We set a different split point as the value computed by an anonymous
function. Here we split on the first quartile along *x*.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.3,
                                          center_x = function(x) {
                                            quantile(x, 
                                                     probs = 1/4, 
                                                     names = FALSE)
                                          },
                                          direction = "split"),
                  min.segment.length = 0)
```

The labels can be rotated as long as the geometry used supports this.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(angle = 90,
                  position = 
                    position_nudge_center(y = 0.1,
                                          direction = "split"))
```

By requesting nudging along *x* and *y* and setting `direction = "split"`
nudging is applied according to the quadrants centred on the centroid of the
data.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.1,
                                          y = 0.15,
                                          direction = "split"))
```

With `direction = "radial"`, the distance nudged away from the center is
the same for all labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.25,
                                          y = 0.4,
                                          direction = "radial"),
                  min.segment.length = 0)
```

As shown above for `direction = "split"` we can set the coordinates of
the center also with `direction = "radial"`.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_s(position = 
                    position_nudge_center(x = 0.125,
                                          y = 0.2,
                                          center_x = -0.5,
                                          direction = "radial"),
                  min.segment.length = 0) +
  expand_limits(x = c(-2.7, +2.3))
```

We can also set the justification of the text labels although repulsion
usually works best with labels justified at the centre, which is the
default in `geom_text_repel()`.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.125,
                                          y = 0.25,
                                          center_x = 0,
                                          center_y = 0,
                                          direction = "radial"),
                  min.segment.length = 0,
                  hjust = "outward", vjust = "outward") +
  expand_limits(x = c(-2.7, +2.3))
```

Nudging along one axis, here *x*, and setting the repulsion `direction`
along the other axis, here *y*, tends to give a pleasant arrangement of
labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text_repel(position = 
                    position_nudge_center(x = 0.2,
                                          center_x = 0,
                                          direction = "split"),
                  aes(hjust = ifelse(x < 0, 1, 0)),
                  direction = "y",
                  min.segment.length = 0) +
  expand_limits(x = c(-3, 3))
```

### position_nudge_line()

Function `position_nudge_line()` nudges away from a line, which can be a
user supplied straight line as well as a smooth spline or a polynomial
fitted to the observations themselves. The nudging is away and
perpendicular to the local slope of the straight or curved line. It
relies on the same assumptions as linear regression, assuming that *x*
values are not subject to error. This in most cases prevents labels from
overlaping a curve fitted to the data, even if not exactly based on the
same model fit. When observations are sparse, this may be enough to
obtain a nice arrangement of data labels.

```{r}
set.seed(16532)
df <- data.frame(
  x = -10:10,
  y = (-10:10)^2,
  yy = (-10:10)^2 + rnorm(21, 0, 4),
  yyy = (-10:10) + rnorm(21, 0, 4),
  l = letters[1:21]
)
```

The first, simple example shows that `position_nudge_line()` has shifted
the direction of the nudging based on the alignment of the observations
along a line. One could, of course, have in this case passed suitable
values as arguments to *x* and *y* using `position_nudge()` from package
'ggplot2'. However, `position_nudge_line()` will work without change
irrespective of the slope or intercept along which the observations
fall.

```{r}
ggplot(df, aes(x, 2 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 2, linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.5, y = -0.8))
```

With observations with high variation in *y*, a linear model fit may
need to be used. In this case fitted twice, once in `stat_smooth()` and
once in `position_nudge_line()`.

```{r}
ggplot(subset(df, x >= 0), aes(x, yyy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            vjust = "center", hjust = "center",
            position = position_nudge_line(x = 0, y = 1.2,
                                           method = "lm",
                                           direction = "split"))
```

With lower variation in *y*, we can pass to `line_nudge` a multiplier to
keep labels outside of the confidence band.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            position = position_nudge_line(method = "lm",
                                           x = 3, y = 3, 
                                           line_nudge = 2.5,
                                           direction = "split"))
```

If we want the nudging based on an arbitrary straight line not computed
from `data`, we can pass the intercept and slope in a numeric vector of
length two as an argument to parameter `abline`.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  geom_text(aes(label = l),
            position = position_nudge_line(abline = c(0, 1),
                                           x = 3, y = 3, 
                                           direction = "split"))
```

With observations that follow exactly a simple curve the defaults work
well to automate the nudging of individual data labels away from the
implicit curve. Positive values as arguments to `x` and `y` correspond
to above and inside the curve. One could, of course, pass also in this
case suitable values as arguments to *x* and *y* using
`position_nudge()` from package 'ggplot2', but these arguments would
need to be vectors with different *nudge* values for each observation.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = 0.6, y = 6))
```

Negative values passed as arguments to `x` and `y` correspond to labels
below and outside the curve.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.6, y = -6))
```

When the observations include random variation along *y*, it is
important that the smoother used for the line added to a plot and that
passed to `position_nudge_line()` are similar. By default
`stat_smooth()` uses `"loess"` and `position_nudge_line()` with method
`"spline"`, `smooth.sline()`, which are a good enough match.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(x = 0.6, 
                                           y = 6,
                                           direction = "split"))
```

We can use other geometries, or rather we need to use a repulsive
geometry when the label text is long or the labels are crowed near the
line.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
                   position = position_nudge_line(x = 0.6, 
                                                  y = 8,
                                                  direction = "split"),
                   box.padding = 0.2,
                   min.segment.length = 0)
```

We can see by comparing the plot above with that below, that combining
nudging away from a line with repulsion results in a more pleasant
positioning of the data labels.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_label_repel(aes(y = yy, label = paste("point", l)),
                  box.padding = 0.5,
                  min.segment.length = 0)
```

When fitting a polynomial, `"lm"` should be the argument passed to
`method` and a model formula preferably based on `poly()`, setting
`raw = TRUE`, as argument to `formula`.

*Currently no other methods are implemented in* `position_nudge_line()`.

```{r}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(method = "lm",
                                           formula = y ~ poly(x, 2, raw = TRUE),
                                           x = 0.5, 
                                           y = 5,
                                           direction = "split"))
```

Compared to pure repulsion adding nudging helps "direct" repulsion in
the desired direction. Compare the plot below using no nudging with the
one above.

```{r, eval=eval_ggrepel}
ggplot(df, aes(x, yy)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text_repel(aes(y = yy, label = l),
                  box.padding = 0.5,
                  min.segment.length = Inf)
```

## Combined position functions

Using `position_stacknudge()` together `geom_label_repel()` makes it
possible to use repulsion for labeling sections of stacked column plots.

```{r, eval=eval_ggrepel}
df <- tibble::tribble(
  ~y, ~x, ~grp,
  "a", 1,  "some long name",
  "a", 2,  "other name",
  "b", 1,  "some name",
  "b", 3,  "another name",
  "b", -1, "some long name"
)

ggplot(data = df, aes(x, y, group = grp)) +
  geom_col(aes(fill = grp), width=0.5) +
  geom_vline(xintercept = 0) +
  geom_label_repel(aes(label = grp),
                   position = position_stacknudge(vjust = 0.5, y = 0.4),
                   label.size = NA)
```

## Acknowledgements

I warmly thank Kamil Slowikowski for agreeing to make changes in
'ggrepel' that make the use of 'ggrepel' together with 'ggpp' possible
and smooth. This document shows some use examples, but surely new ones
will be found by users of R and 'ggplot2'.
