---
title: "How to add padding"
output: html_document
date: "2022-09-26"
---

(https://stackoverflow.com/questions/22649781/shorten-arrows-lines-segments-between-coordinates)

These are all great answers. In an attempt to come up with something that doesn't presume that points connect in a chain, I wrote the following function, which moves x0y0 (a dataframe where column 1 is x and column 2 is y) closer to xy (same format as x0y0) by absolute distance d.

```{r}
movePoints <- function(x0y0, xy, d){
  total.dist <- apply(cbind(x0y0, xy), 1,
             function(x) stats::dist(rbind(x[1:2], x[3:4])))
  p <- d / total.dist
  p <- 1 - p
  x0y0[,1] <- xy[,1] + p*(x0y0[,1] - xy[,1])
  x0y0[,2] <- xy[,2] + p*(x0y0[,2] - xy[,2])
  return(x0y0)
}
```

Possible implementation of segment shrinking as a fraction of starting length.

```{r}
shorten_segments <- function(data, 
                             label.padding = 0,
                             point.padding = 0,
                             min.segment.length = 0.5) {
  stopifnot(label.padding >= 0 & point.padding >= 0 & (label.padding + point.padding) < 1)
  segments.data <- data[ , c("x_orig", "y_orig", "x", "y")]
  starting.length <- apply(segments.data, 1,
                           function(x) stats::dist(rbind(x[1:2], x[3:4])))
  # padding origin
  if (point.padding != 0) {
    p <- point.padding / starting.dist
    p <- 1 - p
    segments.data[ , "x_orig"] <- data[ , "x"] + p * (data[ ,"x_orig"] - data[ , "x"])
    segments.data[ , "y_orig"] <- data[ , "y"] + p * (data[ ,"y_orig"] - data[ , "y"])
  }
  # padding position
  if (label.padding != 0) {
    p <- -label.padding / starting.dist
    p <- 1 - p
    segments.data[ , "x"] <- data[ , "x_orig"] + p * (data[ ,"x"] - data[ , "x_orig"])
    segments.data[ , "y"] <- data[ , "y_orig"] + p * (data[ ,"y"] - data[ , "y_orig"])
  }
  final.length <- apply(segments.data, 1,
                        function(x) stats::dist(rbind(x[1:2], x[3:4])))
  segments.data$too.short <- final.length < min.segment.length
  segments.data
}
```

```{r}
data <- data.frame(x_orig = c(2:3, -1), y_orig = 0, x = 1:3, y = 0)
shorten_segments(data, point.padding = 0.01, label.padding = 0.1, 
                 min.segment.length = 0)
data
```

movePoints <- function(x0y0, xy, d){
  total.dist <- apply(cbind(x0y0, xy), 1,
             function(x) stats::dist(rbind(x[1:2], x[3:4])))
  p <- d / total.dist
  p <- 1 - p
  x0y0[,1] <- xy[,1] + p*(x0y0[,1] - xy[,1])
  x0y0[,2] <- xy[,2] + p*(x0y0[,2] - xy[,2])
  return(x0y0)
}
