---
title: "Position functions"
subtitle: "Advanced nudging"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Position functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
```

## Preliminaries

As we will use text and labels on the plotting area we change the
default theme to an uncluttered one.

```{r}
library(ggpmisc)
old_theme <- theme_set(theme_bw())
```

## Position functions

Function `position_nudge()` from package 'ggplot2' simply applies the
nudge, or *x* and *y* shift based directly on the values passed to its
parameters `nudge_x` and `nudge_y`. In some cases, this means manually
adjusting the nudge based on the plotted data.

We can do better, by setting a point or line away from which the nudge
should be applied. In the new functions this reference alters only the
direction (angle) along which nudge is applied but not the distance.

### position_nudge_center()

Function `position_nudge_center()` limits the directions to radiate away
from a reference point if both *nudge_x* and *nudge_y* are passed an
argument, or if only one of *nudge_x* or *nudge_y* is passed an
argument, nudge right and left, or nudge up and down, respectively. By
default, the "center" is the centroid computed using `mean()`, but other
functions or numeric values can be passed to override it. When data are
sparse, such nudging may be effective in avoiding label overlaps, and
achieving a visually pleasing posiitoning.

```{r}
set.seed(84532)
df <- data.frame(
  x = rnorm(20),
  y = rnorm(20, 2, 2),
  l = sample(letters[1:6], 20, replace = TRUE)
)

```

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1))
```
By default, split based on `mean()`.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1,
                                             direction = "split"))
```

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1,
                                             center_x = 1,
                                             direction = "split"))
```

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1,
                                             y = 0.15,
                                             direction = "split"))
```

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1,
                                             y = 0.3,
                                             direction = "radial"))
```

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_text(position = position_nudge_center(x = 0.1,
                                             y = 0.3,
                                             center_x = 0,
                                             center_y = 0,
                                             direction = "radial"))
```

### position_nudge_line()

Function `position_nudge_line()` nudges away from a line, which can be a
user supplied straight line as well as a smooth spline or a polynomial
fitted to the observations themselves. The nudging is away and
perpendicular to the local slope of the straight or curved line. It
relies on the same assumptions as linear regression, assuming that *x*
values are not subject to error. This in most cases prevents labels from
overlaping a curve fitted to the data, even if not exactly based on the
same model fit. When observations are sparse, this may be enough to
obtain a nice arrangement of data labels.

```{r}
set.seed(16532)
df <- data.frame(
  x = -10:10,
  y = (-10:10)^2,
  yy = (-10:10)^2 + rnorm(21, 0, 4),
  yyy = (-10:10) + rnorm(21, 0, 4),
  l = letters[1:21]
)
```

The first, simple example shows that `position_nudge_line()` has shifted
the direction of the nudging based on the alignment of the observations
along a line. One could, of course, have in this case passed suitable
values as arguments to *x* and *y* using `position_nudge()` from package
'ggplot2'.

Default in 'ggplot2' is no nudging.

```{r}
ggplot(df, aes(x, 2 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 2, linetype = "dotted") +
  geom_text(position = position_nudge(x = 0.4, y = -0.7))
```
Same for `position_nudge_line()`.

```{r}
ggplot(df, aes(x, 2 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 2, linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.5, y = -0.8))
```

However, `position_nudge_line()` will work without change irrespective
of the slope or intercept along which the observations fall.

```{r}
ggplot(df, aes(x, -3 * x, label = l)) +
  geom_point() +
  geom_abline(intercept = 0, slope = -3, linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.5, y = -0.8))
```

With observations with high variation in *y*, a linear model fit may
need to be used.

```{r}
ggplot(subset(df, x >= 0), aes(x, yyy)) +
  geom_point() +
  stat_smooth(method = "lm") +
  geom_text(aes(label = l),
            vjust = "center", hjust = "center",
            position = position_nudge_line(x = 0, y = 1.2,
                                           method = "lm",
                                           direction = "split"))
```

With lower variation in *y*, we can pass to `line_nudge` a multiplier to
keep labels outside of the confidence band.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  geom_text(aes(label = l),
            position = position_nudge_line(method = "lm",
                                           x = 3, y = 3, 
                                           line_nudge = 2.5,
                                           direction = "split"))
```

If we want the nudging based on an arbitrary straight line not computed
from `data`, we can pass the intercept and slope in a numeric vector of
length two as an argument to parameter `abline`.

```{r}
ggplot(subset(df, x >= 0), aes(y, yy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  geom_text(aes(label = l),
            position = position_nudge_line(abline = c(0, 1),
                                           x = 3, y = 3, 
                                           direction = "split"))
```

With observations that follow exactly a simple curve the defaults work well to
automate the nudging of individual data labels away from the implicit curve.
Positive values as arguments to `x` and `y` correspond to above and inside the
curve. One could, of course, pass also in this case suitable values as arguments
to *x* and *y* using `position_nudge()` from package 'ggplot2', but these
arguments would need to be vectors with different *nudge* values for each
observation.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = 0.6, y = 6))
```

Negative values passed as arguments to `x` and `y` correspond to labels
below and outside the curve.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_point() +
  geom_line(linetype = "dotted") +
  geom_text(position = position_nudge_line(x = -0.6, y = -6))
```

When the observations include random variation along *y*, it is
important that the smoother used for the line added to a plot and that
passed to `position_nudge_line()` are similar. By default
`stat_smooth()` uses `"loess"` and `position_nudge_line()` with methid
`"spline"`, `smooth.sline()`, which are a good enough match.

```{r}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth() +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(x = 0.6, 
                                           y = 6,
                                           direction = "split"))
```

When fitting a polynomial, `"lm"` should be the argument passed to
`method` and a model formula preferably based on `poly()`, setting `raw = TRUE`,
as argument to `formula`.
*Currently no other methods are implemented in* `position_nudge_line()`.

```{r}
ggplot(df, aes(x, yy, label = l)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE)) +
  geom_text(aes(y = yy, label = l),
            position = position_nudge_line(method = "lm",
                                           x = 0.6, 
                                           y = 6,
                                           formula = y ~ poly(x, 2, raw = TRUE),
                                           direction = "split"))
```

Grouping is supported.

```{r}
df <- data.frame(x = rep(1:10, 2),
                 y = c(1:10, 10:1),
                 group = rep(c("a", "b"), c(10, 10)),
                 l = "+")
```

```{r}
ggplot(df, aes(x, y, label = l, color = group)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(position = position_nudge_line(x = 0.25, y = 1)) +
  geom_text(position = position_nudge_line(x = -0.25, y = -1)) +
  coord_equal(ratio = 0.5)
```

One needs to ensure that grouping is in effect in the geoms with nudging.

```{r}
ggplot(df, aes(x, y, label = l, color = group, group = group)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(color = "red",
            position = position_nudge_line(x = 1, y = 1)) +
  geom_text(color = "blue",
            position = position_nudge_line(x = -1, y = -1)) +
  coord_equal()
```

Facets are also supported.

```{r}
ggplot(df, aes(x, y, label = l)) +
  geom_line(linetype = "dotted") +
  geom_text() +
  geom_text(position = position_nudge_line(x = 1, y = 1),
            color = "red") +
  geom_text(position = position_nudge_line(x = -1, y = -1),
            color = "blue") +
  facet_wrap(~group) +
  coord_equal(ratio = 1)
```

